---
title: "Data Visualization - day2"
subtitle: "<span style='font-size:1.25em;'>for Genomics and Biological Sciences</span>"
author: "<div style='width: 100%; text-align: center; margin-left: auto; margin-right: auto;'><a href='https://github.com/loukesio' style='color:#28a87d;font-weight:600;'>Dr. Loukas Theodosiou</a>&ensp;//&ensp;December 15<sup>th</sup> 2025</div>"
title-slide-attributes:
    data-background-color: "#E8F5F1"
format:
  revealjs:
    theme: slides.scss
    footer: '**L.Theodosiou** // Consultation in Genomics Data Visualization'
    transition: fade
    progress: true
    multiplex: false
    preview-links: true
    scrollable: false
    hide-inactive-cursor: true
    highlight-style: printing
    pause: true
    embed-resources: false
editor: source
knitr:
  opts_chunk:
    dev: "ragg_png"
    retina: 3
    dpi: 200
execute:
  freeze: auto
  cache: true
  echo: true
  fig-width: 12
  fig-height: 7

---


```{r}
#| label: setup
#| include: false

library(tidyverse)      # Includes dplyr, ggplot2, tidyr, readr
library(patchwork)      # Combine multiple plots

# Color palettes and tables
library(MetBrewer)      # Color palettes
library(RColorBrewer)   # Additional color palettes
library(gt)             # Publication-quality tables

# Specialized ggplot extensions
library(ggrepel)        # Non-overlapping text labels
library(ggforce)        # Additional ggplot2 features

# Volcano plots
library(ggvolc)         # Genomics volcano plots
library(EnhancedVolcano)  # Bioconductor volcano plots

# Heatmaps
library(pheatmap)       # Pretty heatmaps
library(ComplexHeatmap) # Advanced heatmaps (Bioconductor)

# Set comparisons
library(VennDiagram)    # Venn diagrams
library(UpSetR)         # UpSet plots for complex overlaps

# Dimensionality reduction
library(umap)           # UMAP
library(Rtsne)          # t-SNE

# Genomics visualization
library(GenomicRanges)  # Genomic intervals (Bioconductor)
library(IRanges)        # Integer ranges (Bioconductor)
library(karyoploteR)    # Karyotype plots (Bioconductor)
library(Gviz)           # Genomic tracks (Bioconductor)

# Network and graph visualization
library(tidygraph)      # Tidy graph manipulation
library(igraph)         # Network analysis
library(ggraph)         # Network visualization with ggplot2

# Interactive plots
library(plotly)         # Interactive plots


# Define a custom theme function for ggplot2 to enhance plot aesthetics
custom_theme <- function(base_size = 12, title_size_rel = 1.5,
                         subtitle_size_rel = 1.25,
                         axis_title_size_rel = 1.25, 
                         axis_text_size_rel = 1.25, 
                         strip_text_size_rel = 1.1) {
  # Use the theme_bw base theme with a specified font family ("Asap")
  theme_bw(base_size = base_size, base_family = "Asap") +
    theme(
      # Customize the title: center-align and set relative size
      plot.title = element_text(hjust = 0.5, size = rel(title_size_rel)),
      # Customize the subtitle: center-align and set relative size
      plot.subtitle = element_text(hjust = 0.5, size = rel(subtitle_size_rel)),
      # Customize the x-axis title: set size and add a top margin
      axis.title.x = element_text(size = rel(axis_title_size_rel), 
                                  margin = margin(t = 12.5)),
      # Customize the y-axis title: set size and add a right margin
      axis.title.y = element_text(size = rel(axis_title_size_rel), 
                                  margin = margin(r = 12.5)),
      # Customize the x-axis text: set size
      axis.text.x = element_text(size = rel(axis_text_size_rel)),
      # Customize the y-axis text: set size
      axis.text.y = element_text(size = rel(axis_text_size_rel)),
      # Customize facet strip text size
      strip.text = element_text(size = rel(strip_text_size_rel)),
      # Add margins around the plot t,r,b,l
      plot.margin = unit(c(2, 2, 1, 1), "cm"),
      # Remove major grid lines on the x-axis
      panel.grid.major.x = element_blank(),
      panel.grid.major.y  = element_blank(),
      # Remove minor grid lines
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_line(
      linetype = "dashed",
      linewidth = 0.3,
      colour = "grey80"
      ),      # Remove background from facet strips
      strip.background = element_blank()
    )
}

theme_set(custom_theme())


cross_colors <- met.brewer("Cross", 5)
```

## Why Reduce Dimensions?

**High-dimensional data challenges:**

- Genomic datasets often have thousands of features
- Difficult to visualize and interpret
- Computationally expensive

::: fragment
**Solution: Dimensionality reduction**

- Compress data to 2-3 dimensions
- Preserve important patterns and relationships
- Enable visual exploration
:::
```{=html}
<!-- slide 36 -->
```

## PCA: Theory

**Principal Component Analysis**

- **Linear method** that finds directions of maximum variance
- Creates orthogonal (uncorrelated) components
- PC1 explains most variance, PC2 second most, etc.

::: fragment
**Best for:**

- Linearly separable data
- Understanding variance structure
- Quality control and exploratory analysis
- Fast computation on large datasets
:::

::: fragment
**Key concept:** Transforms correlated variables into uncorrelated principal components
:::
```{=html}
<!-- slide 37 -->
```

## PCA: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-8|10-16"

# Create simulated data with 3 groups
set.seed(123)
pca_demo <- data.frame(
  PC1 = c(rnorm(100, -3, 1), rnorm(100, 0, 1), rnorm(100, 3, 1)),
  PC2 = c(rnorm(100, 2, 1), rnorm(100, 0, 1), rnorm(100, -2, 1)),
  Group = rep(c("Group A", "Group B", "Group C"), each = 100)
)

# Visualize with ellipses
ggplot(pca_demo, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 2.5, alpha = 0.6) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(title = "PCA: Linear Separation",
       x = "PC1 (45.2%)", y = "PC2 (23.8%)") +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 37a -->
```

## PCA: Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

set.seed(123)
pca_demo <- data.frame(
  PC1 = c(rnorm(100, -3, 1), rnorm(100, 0, 1), rnorm(100, 3, 1)),
  PC2 = c(rnorm(100, 2, 1), rnorm(100, 0, 1), rnorm(100, -2, 1)),
  Group = rep(c("Group A", "Group B", "Group C"), each = 100)
)

ggplot(pca_demo, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 2.5, alpha = 0.6) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(title = "PCA: Linear Separation",
       x = "PC1 (45.2%)", y = "PC2 (23.8%)") +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 37b -->
```

## UMAP: Theory

**Uniform Manifold Approximation and Projection**

- **Non-linear method** based on manifold learning
- Preserves both local and global structure
- Faster than t-SNE for large datasets

::: fragment
**Best for:**

- Complex, non-linear relationships
- Preserving cluster relationships
- Large datasets (>10,000 cells/samples)
- General exploratory analysis
:::

::: fragment
**Key concept:** Models data as a manifold and projects to lower dimensions while preserving topology
:::
```{=html}
<!-- slide 38 -->
```

## UMAP: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-11|13-18"

# Simulate UMAP-like structure
set.seed(456)
theta <- seq(0, 2*pi, length.out = 100)
umap_demo <- data.frame(
  UMAP1 = c(2*cos(theta) + rnorm(100, 0, 0.3),
            5 + 1.5*cos(theta) + rnorm(100, 0, 0.3),
            2.5 + 1.5*sin(theta*1.5) + rnorm(100, 0, 0.3)),
  UMAP2 = c(2*sin(theta) + rnorm(100, 0, 0.3),
            1.5*sin(theta) + rnorm(100, 0, 0.3),
            5 + 1.5*cos(theta*1.5) + rnorm(100, 0, 0.3)),
  Group = rep(c("Group A", "Group B", "Group C"), each = 100)
)

# Visualize
ggplot(umap_demo, aes(x = UMAP1, y = UMAP2, color = Group)) +
  geom_point(size = 2.5, alpha = 0.6) +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(title = "UMAP: Preserves Local & Global Structure")
```
```{=html}
<!-- slide 38a -->
```

## UMAP: Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center
#| message: false

set.seed(456)
theta <- seq(0, 2*pi, length.out = 100)
umap_demo <- data.frame(
  UMAP1 = c(2*cos(theta) + rnorm(100, 0, 0.3),
            5 + 1.5*cos(theta) + rnorm(100, 0, 0.3),
            2.5 + 1.5*sin(theta*1.5) + rnorm(100, 0, 0.3)),
  UMAP2 = c(2*sin(theta) + rnorm(100, 0, 0.3),
            1.5*sin(theta) + rnorm(100, 0, 0.3),
            5 + 1.5*cos(theta*1.5) + rnorm(100, 0, 0.3)),
  Group = rep(c("Group A", "Group B", "Group C"), each = 100)
)

ggplot(umap_demo, aes(x = UMAP1, y = UMAP2, color = Group)) +
  geom_point(size = 2.5, alpha = 0.6) +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(title = "UMAP: Preserves Local & Global Structure",
       x = "UMAP1", y = "UMAP2") +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 38b -->
```

## t-SNE: Theory

**t-Distributed Stochastic Neighbor Embedding**

- **Non-linear method** focusing on local structure
- Converts high-dimensional distances to probabilities
- Excellent at revealing tight clusters

::: fragment
**Best for:**

- Cluster visualization and discovery
- Revealing hidden structure
- Smaller datasets (<10,000 points)
- Publications requiring clear separation
:::

::: fragment
**Caution:** 

- Distances between clusters are not meaningful
- Different runs produce different layouts (non-deterministic)
:::
```{=html}
<!-- slide 39 -->
```

## t-SNE: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-7|9-14"

# Simulate t-SNE-like structure
set.seed(789)
tsne_demo <- data.frame(
  tSNE1 = c(rnorm(100, -5, 0.5), rnorm(100, 0, 0.5), rnorm(100, 5, 0.5)),
  tSNE2 = c(rnorm(100, 3, 0.5), rnorm(100, -3, 0.5), rnorm(100, 0, 0.5)),
  Group = rep(c("Group A", "Group B", "Group C"), each = 100)
)

# Visualize
ggplot(tsne_demo, aes(x = tSNE1, y = tSNE2, color = Group)) +
  geom_point(size = 2.5, alpha = 0.6) +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(title = "t-SNE: Emphasizes Local Clusters",
       x = "t-SNE1", y = "t-SNE2")
```
```{=html}
<!-- slide 39a -->
```

## t-SNE: Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

set.seed(789)
tsne_demo <- data.frame(
  tSNE1 = c(rnorm(100, -5, 0.5), rnorm(100, 0, 0.5), rnorm(100, 5, 0.5)),
  tSNE2 = c(rnorm(100, 3, 0.5), rnorm(100, -3, 0.5), rnorm(100, 0, 0.5)),
  Group = rep(c("Group A", "Group B", "Group C"), each = 100)
)

ggplot(tsne_demo, aes(x = tSNE1, y = tSNE2, color = Group)) +
  geom_point(size = 2.5, alpha = 0.6) +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(title = "t-SNE: Emphasizes Local Clusters",
       x = "t-SNE1", y = "t-SNE2") +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 39b -->
```

## Choosing the Right Method
```{r}
#| echo: false

comparison_table <- tibble(
  Method = c("PCA", "UMAP", "t-SNE"),
  Type = c("Linear", "Non-linear", "Non-linear"),
  Speed = c("Fast", "Medium-Fast", "Slow"),
  `Preserves` = c(
    "Global variance",
    "Local & global structure",
    "Local neighborhoods"
  ),
  `Best For` = c(
    "QC, linear relationships",
    "General exploration, large data",
    "Cluster discovery, small data"
  ),
  Deterministic = c("Yes", "No (but reproducible)", "No")
)

comparison_table |>
  gt() |>
  cols_align(align = "left") |>
  tab_style(
    style = cell_text(weight = "bold", size = px(14)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(size = px(12)),
    locations = cells_body()
  ) |>
  tab_options(
    table.font.size = px(13),
    table.width = pct(95)
  )
```
```{=html}
<!-- slide 40 -->
```

```{=html}
<!-- slide 40 -->
```

## PCA with prcomp(): Code
```{r}
#| eval: false
#| code-line-numbers: "|1-10|12-15|17-20|22-29"

# Simulate gene expression matrix (samples × genes)
set.seed(42)
n_samples <- 60
n_genes <- 500

expr_data <- matrix(
  rnorm(n_samples * n_genes, mean = 50, sd = 20),
  nrow = n_samples,
  ncol = n_genes
)

# Add sample metadata
sample_info <- data.frame(
  sample_id = paste0("Sample_", 1:n_samples),
  tissue = rep(c("Brain", "Liver", "Heart"), each = 20)
)

# Perform PCA (samples as rows)
pca_result <- prcomp(expr_data, scale. = TRUE)
var_exp <- summary(pca_result)$importance[2,]

# Extract PC scores and add metadata
pca_scores <- pca_result$x |> 
  as.data.frame() |> 
  bind_cols(sample_info)

# Plot
ggplot(pca_scores, aes(x = PC1, y = PC2, color = tissue)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse() +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(
    title = "PCA of Gene Expression",
    x = paste0("PC1 (", round(var_exp[1]*100, 1), "%)"),
    y = paste0("PC2 (", round(var_exp[2]*100, 1), "%)")
  )
```
```{=html}
<!-- slide 41 -->
```

## PCA with prcomp(): Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

set.seed(42)
n_samples <- 60
n_genes <- 500

# Create expression with tissue-specific patterns
expr_data <- matrix(
  rnorm(n_samples * n_genes, mean = 50, sd = 20),
  nrow = n_samples,
  ncol = n_genes
)

# Add tissue-specific variation
expr_data[1:20, 1:200] <- expr_data[1:20, 1:200] + 30
expr_data[21:40, 201:400] <- expr_data[21:40, 201:400] + 30
expr_data[41:60, 401:500] <- expr_data[41:60, 401:500] + 30

sample_info <- data.frame(
  sample_id = paste0("Sample_", 1:n_samples),
  tissue = rep(c("Brain", "Liver", "Heart"), each = 20)
)

pca_result <- prcomp(expr_data, scale. = TRUE)

pca_scores <- pca_result$x |> 
  as.data.frame() |> 
  bind_cols(sample_info)

var_exp <- summary(pca_result)$importance[2,]

ggplot(pca_scores, 
       aes(x = PC1, y = PC2, color = tissue)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse() +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(
    title = "PCA of Gene Expression",
    x = paste0("PC1 (", round(var_exp[1]*100, 1), "%)"),
    y = paste0("PC2 (", round(var_exp[2]*100, 1), "%)")
  ) +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 42 -->
```

## Exercise 1: PCA on Gene Expression Data
```{r}
#| eval: false
#| code-line-numbers: "|1-10|12-17|19-21|23-31|33-42"

# 1. Create gene expression matrix (samples × genes)
set.seed(123)
n_samples <- 40
n_genes <- 1000

gene_expr <- matrix(
  rpois(n_samples * n_genes, lambda = 50),
  nrow = n_samples,
  ncol = n_genes
)

# Add column names (gene IDs)
colnames(gene_expr) <- paste0("Gene_", 1:n_genes)

# Create sample metadata
sample_meta <- data.frame(
  sample_id = paste0("Sample_", 1:n_samples),
  condition = rep(c("Control", "Treatment"), each = 20),
  batch = rep(1:4, times = 10)
)

# 2. Perform PCA with scaling
pca_result <- prcomp(gene_expr, scale. = TRUE)
summary(pca_result)

# 3. Extract scores and combine with metadata
pca_df <- pca_result$x |> 
  as.data.frame() |> 
  bind_cols(sample_meta)

# 4. Create PCA plot colored by condition
ggplot(pca_df, aes(x = PC1, y = PC2, 
                   color = condition, 
                   shape = factor(batch))) +
  geom_point(size = 4, alpha = 0.7) +
  stat_ellipse(aes(color = condition)) +
  scale_color_manual(values = c("Control" = "steelblue", 
                                 "Treatment" = "coral")) +
  labs(
    title = "PCA of RNA-seq Data",
    subtitle = "Check for batch effects and treatment separation"
  )
```

**Questions:** How much variance is explained by PC1 and PC2? Do you see batch effects? Are treatment groups well separated?
```{=html}
<!-- slide 43 -->
```

## Exercise 2: Scree Plot and Gene Loadings
```{r}
#| eval: false
#| code-line-numbers: "|1-12|14-19|21-30"

# 1. Create scree plot
variance_explained <- summary(pca_result)$importance[2,]

scree_data <- data.frame(
  PC = paste0("PC", 1:length(variance_explained)),
  Variance = variance_explained * 100
)

ggplot(scree_data, aes(x = PC, y = Variance)) +
  geom_col(fill = "steelblue") +
  geom_line(aes(group = 1), color = "#d1422f", linewidth = 1) +
  geom_point(color = "#d1422f", size = 3) +
  labs(title = "Scree Plot: Variance Explained")

# 2. Find top contributing genes to PC1
loadings_pc1 <- pca_result$rotation[, 1]
top_genes <- sort(abs(loadings_pc1), decreasing = TRUE)[1:10]
print("Top 10 genes driving PC1:")
print(top_genes)

# 3. Create loading plot for PC1 vs PC2
loadings_df <- pca_result$rotation[, 1:2] |> 
  as.data.frame() |> 
  mutate(gene = rownames(pca_result$rotation),
         total_loading = sqrt(PC1^2 + PC2^2)) |>
  arrange(desc(total_loading)) |>
  slice_head(n = 20)

ggplot(loadings_df, aes(x = PC1, y = PC2, label = gene)) +
  geom_segment(aes(xend = 0, yend = 0), 
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_text(size = 3, hjust = -0.1) +
  labs(title = "Gene Loadings: PC1 vs PC2")
```

**Question:** Which genes are most important for separating your samples?
```{=html}
<!-- slide 44 -->
```

## Exercise 3: Compare PCA, UMAP, and t-SNE
```{r}
#| eval: false
#| code-line-numbers: "|1-4|6-12|14-20|22-28|30-32|34-43"

# Install packages if needed
# install.packages(c("umap", "Rtsne"))
library(umap)
library(Rtsne)

# 1. Scale the gene expression data
scaled_expr <- scale(gene_expr)

# 2. PCA
pca_res <- prcomp(scaled_expr)
pca_df <- data.frame(
  Dim1 = pca_res$x[,1],
  Dim2 = pca_res$x[,2],
  Method = "PCA",
  condition = sample_meta$condition,
  batch = factor(sample_meta$batch)
)

# 3. UMAP
set.seed(123)
umap_res <- umap(scaled_expr, n_neighbors = 15)
umap_df <- data.frame(
  Dim1 = umap_res$layout[,1],
  Dim2 = umap_res$layout[,2],
  Method = "UMAP",
  condition = sample_meta$condition,
  batch = factor(sample_meta$batch)
)

# 4. t-SNE
set.seed(123)
tsne_res <- Rtsne(scaled_expr, perplexity = 10, check_duplicates = FALSE)
tsne_df <- data.frame(
  Dim1 = tsne_res$Y[,1],
  Dim2 = tsne_res$Y[,2],
  Method = "t-SNE",
  condition = sample_meta$condition,
  batch = factor(sample_meta$batch)
)

# 5. Combine and plot
all_methods <- bind_rows(pca_df, umap_df, tsne_df)

ggplot(all_methods, aes(x = Dim1, y = Dim2, 
                        color = condition, shape = batch)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_wrap(~Method, scales = "free") +
  scale_color_manual(values = c("Control" = "steelblue", 
                                 "Treatment" = "coral")) +
  labs(title = "Comparison: PCA vs UMAP vs t-SNE") +
  theme(legend.position = "bottom")
```

**Questions:** Which method best separates treatment groups? Which shows clearest batch effects?
```{=html}
<!-- slide 45 -->
```

## Exercise 4: Single-Cell RNA-seq Simulation
```{r}
#| eval: false
#| code-line-numbers: "|1-10|12-19|21-29|31-33|35-44|46-50"

# 1. Simulate single-cell RNA-seq data
set.seed(456)
n_cells <- 500
n_genes <- 2000

scrna_expr <- matrix(
  rnorm(n_cells * n_genes, mean = 0, sd = 1),
  nrow = n_cells,
  ncol = n_genes
)

# Add cell type-specific expression patterns
scrna_expr[1:150, 1:500] <- scrna_expr[1:150, 1:500] + 3      # T cells
scrna_expr[151:300, 501:1000] <- scrna_expr[151:300, 501:1000] + 3  # B cells
scrna_expr[301:450, 1001:1500] <- scrna_expr[301:450, 1001:1500] + 3  # Monocytes
scrna_expr[451:500, 1501:2000] <- scrna_expr[451:500, 1501:2000] + 3  # NK cells

# Cell metadata
cell_meta <- data.frame(
  cell_id = paste0("Cell_", 1:n_cells),
  cell_type = c(rep("T cell", 150),
                rep("B cell", 150),
                rep("Monocyte", 150),
                rep("NK cell", 50)),
  donor = rep(paste0("Donor_", 1:5), each = 100)
)

# 2. Perform PCA
pca_sc <- prcomp(scrna_expr, scale. = TRUE)

# 3. Extract PC scores
pca_sc_df <- pca_sc$x |> 
  as.data.frame() |> 
  bind_cols(cell_meta)

var_explained <- summary(pca_sc)$importance[2,]

# 4. Plot PCA colored by cell type
ggplot(pca_sc_df, aes(x = PC1, y = PC2, color = cell_type)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_manual(values = cross_colors[c(1, 2, 3, 5)]) +
  labs(
    title = "PCA of Single-Cell RNA-seq",
    x = paste0("PC1 (", round(var_explained[1]*100, 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2]*100, 1), "%)"),
    color = "Cell Type"
  )

# 5. BONUS: Check for donor effects
ggplot(pca_sc_df, aes(x = PC1, y = PC2, color = donor)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(title = "Check for Donor Batch Effects")
```

**Questions:** Are cell types well separated? Do you see donor effects? How many PCs would you keep?
```{=html}
<!--
 __      __   _                               _       _       
 \ \    / /  | |                             | |     | |      
  \ \  / /__ | | ___ __ _ _ __   ___    _ __ | | ___ | |_ ___ 
   \ \/ / _ \| |/ __/ _` | '_ \ / _ \  | '_ \| |/ _ \| __/ __|
    \  / (_) | | (_| (_| | | | | (_) | | |_) | | (_) | |_\__ \
     \/ \___/|_|\___\__,_|_| |_|\___/  | .__/|_|\___/ \__|___/
                                       | |                    
                                       |_|      
-->
```


## Volcano Plots: Introduction

**What is a volcano plot?**

- Combines statistical significance (p-value) with biological significance (fold change)
- X-axis: log2 fold change (effect size)
- Y-axis: -log10(p-value) (statistical significance)

::: fragment
**Key features:**

- Identifies significantly differentially expressed genes
- Shows both magnitude and significance of changes
- Enables labeling of genes of interest
:::
```{=html}
<!-- slide 48 -->
```

## Volcano Plot Anatomy

**How to interpret the plot:**

- **Top**: Most significant genes (low p-value)
- **Right**: Upregulated genes (positive fold change)
- **Left**: Downregulated genes (negative fold change)
- **Center**: Not significantly changed

::: fragment
**Typical thresholds:**

- Fold change: |log2FC| > 1 (2-fold change)
- P-value: p < 0.05 or adjusted p < 0.05
- Combined: both criteria must be met
:::
```{=html}
<!-- slide 49 -->
```

## Step 1: Create Basic Plot - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-8|10-17"

# Simulate differential expression data
set.seed(123)
n_genes <- 5000

volcano_demo <- data.frame(
  gene = paste0("Gene_", 1:n_genes),
  log2FC = rnorm(n_genes, mean = 0, sd = 2),
  pvalue = rbeta(n_genes, 0.1, 1)
) |>
  mutate(log10p = -log10(pvalue))

head(volcano_demo)

# Create basic plot (no colors yet)
ggplot(volcano_demo, aes(x = log2FC, y = log10p)) +
  geom_point(alpha = 0.5, size = 1.5, color = "gray50") +
  labs(title = "Basic Volcano Plot",
       x = "log2 Fold Change", 
       y = "-log10(p-value)")
```
```{=html}
<!-- slide 50 -->
```

## Step 1: Create Basic Plot - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

set.seed(123)
n_genes <- 5000

volcano_demo <- data.frame(
  gene = paste0("Gene_", 1:n_genes),
  log2FC = rnorm(n_genes, mean = 0, sd = 2),
  pvalue = rbeta(n_genes, 0.1, 1)
) |>
  mutate(log10p = -log10(pvalue))

ggplot(volcano_demo, aes(x = log2FC, y = log10p)) +
  geom_point(alpha = 0.5, size = 1.5, color = "gray50") +
  labs(title = "Basic Volcano Plot",
       x = "log2 Fold Change", 
       y = "-log10(p-value)") 
```
```{=html}
<!-- slide 51 -->
```

## Step 2: Add Color Classification - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-8|10-16"

# Add significance classification
volcano_demo <- volcano_demo |>
  mutate(
    significance = case_when(
      abs(log2FC) > 1 & pvalue < 0.05 ~ "Significant",
      TRUE ~ "Not significant"
    )
  )

# Color by significance
ggplot(volcano_demo, aes(x = log2FC, y = log10p, color = significance)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("Not significant" = "gray70", 
                                 "Significant" = "#d1422f")) +
  labs(title = "Volcano Plot with Colors",
       x = "log2 Fold Change", y = "-log10(p-value)")

```
```{=html}
<!-- slide 52 -->
```

## Step 2: Add Color Classification - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

volcano_demo <- volcano_demo |>
  mutate(
    significance = case_when(
      abs(log2FC) > 1 & pvalue < 0.05 ~ "Significant",
      TRUE ~ "Not significant"
    )
  )

ggplot(volcano_demo, aes(x = log2FC, y = log10p, color = significance)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("Not significant" = "gray70", 
                                 "Significant" = "#d1422f")) +
  labs(title = "Volcano Plot with Colors",
       x = "log2 Fold Change", 
       y = "-log10(p-value)",
       color = NULL) +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 53 -->
```

## Step 3: Add Up/Down Classification - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-8|10-17"

# Classify as Up, Down, or Not significant
volcano_demo <- volcano_demo |>
  mutate(
    status = case_when(
      log2FC > 1 & pvalue < 0.05 ~ "Upregulated",
      log2FC < -1 & pvalue < 0.05 ~ "Downregulated",
      TRUE ~ "Not significant"
    )
  )

# Color by direction
ggplot(volcano_demo, aes(x = log2FC, y = log10p, color = status)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("Upregulated" = "#d1422f", 
                                 "Downregulated" = "#1a5b5b",
                                 "Not significant" = "gray70")) +
  labs(title = "Up vs Down Regulation", x = "log2 Fold Change", y = "-log10(p-value)")

```
```{=html}
<!-- slide 54 -->
```

## Step 3: Add Up/Down Classification - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

volcano_demo <- volcano_demo |>
  mutate(
    status = case_when(
      log2FC > 1 & pvalue < 0.05 ~ "Upregulated",
      log2FC < -1 & pvalue < 0.05 ~ "Downregulated",
      TRUE ~ "Not significant"
    )
  )

ggplot(volcano_demo, aes(x = log2FC, y = log10p, color = status)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("Upregulated" = "#d1422f", 
                                 "Downregulated" = "#1a5b5b",
                                 "Not significant" = "gray70")) +
  labs(title = "Up vs Down Regulation", 
       x = "log2 Fold Change", 
       y = "-log10(p-value)",
       color = NULL) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 55 -->
```

## Step 4: Add Threshold Lines - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-8|10-12"

# Add threshold lines
ggplot(volcano_demo, aes(x = log2FC, y = log10p, color = status)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("Upregulated" = "#d1422f", 
                                 "Downregulated" = "#1a5b5b",
                                 "Not significant" = "gray70")) +
  labs(title = "Volcano Plot with Thresholds",
       x = "log2 Fold Change", y = "-log10(p-value)")

# Add dashed lines for thresholds
geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black")
```
```{=html}
<!-- slide 56 -->
```

## Step 4: Add Threshold Lines - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

ggplot(volcano_demo, aes(x = log2FC, y = log10p, color = status)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Upregulated" = "#d1422f", 
                                 "Downregulated" = "#1a5b5b",
                                 "Not significant" = "gray70")) +
  labs(title = "Volcano Plot with Thresholds",
       x = "log2 Fold Change", 
       y = "-log10(p-value)",
       color = NULL) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 57 -->
```

## Step 5: Add Gene Labels - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-2|4-7|9-20"

# Load ggrepel
library(ggrepel)

# Select top 10 genes to label
top_genes <- volcano_demo |>
  filter(status != "Not significant") |>
  arrange(pvalue) |>
  slice_head(n = 10)

# Add labels with ggrepel
ggplot(volcano_demo, aes(x = log2FC, y = log10p, color = status)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_point(data = top_genes, size = 3, alpha = 1) +
  geom_text_repel(data = top_genes, aes(label = gene), size = 3) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  scale_color_manual(values = c("Upregulated" = "#d1422f", 
                                 "Downregulated" = "#1a5b5b",
                                 "Not significant" = "gray70")) +
  labs(title = "Volcano Plot with Gene Labels", x = "log2 Fold Change", y = "-log10(p-value)")
```
```{=html}
<!-- slide 58 -->
```

## Step 5: Add Gene Labels - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center
#| warning: false

library(ggrepel)

top_genes <- volcano_demo |>
  filter(status != "Not significant") |>
  arrange(pvalue) |>
  slice_head(n = 10)

ggplot(volcano_demo, aes(x = log2FC, y = log10p, color = status)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_point(data = top_genes, size = 3, alpha = 1) +
  geom_text_repel(data = top_genes, aes(label = gene), size = 3, max.overlaps = 20) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  scale_color_manual(values = c("Upregulated" = "#d1422f", 
                                 "Downregulated" = "#1a5b5b",
                                 "Not significant" = "gray70")) +
  labs(title = "Volcano Plot with Gene Labels", 
       x = "log2 Fold Change", 
       y = "-log10(p-value)",
       color = NULL) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 59 -->
```

## Method 2: EnhancedVolcano - Introduction

**Bioconductor package with automatic features**

- Automatic gene labeling
- Customizable thresholds
- Publication-ready output
- Rich color schemes

**Installation:**
```r
BiocManager::install("EnhancedVolcano")
library(EnhancedVolcano)
```
```{=html}
<!-- slide 60 -->
```

## EnhancedVolcano: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-2|4-12"

# Prepare data for EnhancedVolcano
library(EnhancedVolcano)

# Create enhanced volcano plot
EnhancedVolcano(volcano_demo,
                lab = volcano_demo$gene,
                x = 'log2FC',
                y = 'pvalue',
                pCutoff = 0.05,
                FCcutoff = 1,
                title = 'EnhancedVolcano Example',
                pointSize = 2, 
                labSize = 3)
```

**Note:** Requires Bioconductor installation
```{=html}
<!-- slide 60 -->
```

## EnhancedVolcano: Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center
#| eval: false

library(EnhancedVolcano)

EnhancedVolcano(volcano_demo,
                lab = volcano_demo$gene,
                x = 'log2FC',
                y = 'pvalue',
                pCutoff = 0.05,
                FCcutoff = 1,
                title = 'EnhancedVolcano Example',
                pointSize = 2, 
                labSize = 3)
```

**Note:** This demonstrates EnhancedVolcano's automatic labeling and styling
```{=html}
<!-- slide 61 -->
```

## Method 3: ggvolc - Introduction

**Specialized genomics volcano plot package**

Created by Dr. Loukas Theodosiou

- Genomics-optimized defaults
- Easy gene highlighting
- Significance segments
- Clean styling

**Installation:**
```r
devtools::install_github("loukesio/ggvolc")
library(ggvolc)
```
```{=html}
<!-- slide 62 -->
```

## ggvolc: Data Format
```{r}
#| eval: false

# Load package and example data
library(ggvolc)
data(all_genes)  # Main DEG results

head(all_genes, 3)
#   genes   baseMean log2FoldChange  lfcSE    stat    pvalue      padj
#   GCR1    7201.58       2.24      0.20    11.19   4.4e-29  2.2e-25
#   OPI10   1009.42      -2.26      0.21   -10.77   4.9e-27  1.2e-23
#   AGA2     249.12       3.83      0.36    10.57   4.1e-26  6.7e-23

# Optional: genes to highlight
data(attention_genes)
head(attention_genes, 3)
#   genes   baseMean log2FoldChange  lfcSE    stat    pvalue      padj
#   THI13    480.52       1.59      0.52    3.04    2.4e-03  1.4e-02
#   FBP1   22710.84      -2.37      0.35   -6.70    2.1e-11  3.3e-09
```

**Required columns:** genes, log2FoldChange, pvalue, padj
```{=html}
<!-- slide 63 -->
```

## ggvolc: Basic Plot - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-3|5-7|9-12"

# Load package
library(ggvolc)
data(all_genes)

# Basic plot without segments
ggvolc(all_genes, 
       add_seg = FALSE)

# Add significance segments
ggvolc(all_genes, 
       add_seg = TRUE) +
  labs(title = "ggvolc with Significance Segments")

```
```{=html}
<!-- slide 64 -->
```

## ggvolc: Basic Plot - Output
```{r}
#| echo: true
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center
#| eval: false

library(ggvolc)
data(all_genes)

ggvolc(all_genes, add_seg = TRUE) +
  labs(title = "ggvolc with Significance Segments") +
  theme(text = element_text(size = 16))
```

**Note:** This shows ggvolc's default styling with automatic thresholds
```{=html}
<!-- slide 65 -->
```

## ggvolc: Highlighting Genes - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-2|4-7|9-13"

# Load data with genes to highlight
data(attention_genes)

# Highlight specific genes
ggvolc(all_genes, 
       attention_genes,  # Genes to highlight
       add_seg = TRUE)

# Customize with ggplot2 layers
ggvolc(all_genes, attention_genes, add_seg = TRUE) +
  labs(title = "Highlighted Genes of Interest",
       subtitle = "Using ggvolc with attention_genes") +
  theme_minimal(base_size = 14)
```
```{=html}
<!-- slide 66 -->
```

## ggvolc: Highlighting Genes - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center
#| eval: false

library(ggvolc)
data(all_genes)
data(attention_genes)

ggvolc(all_genes, attention_genes, add_seg = TRUE) +
  labs(title = "Highlighted Genes of Interest",
       subtitle = "Using ggvolc with attention_genes") +
  theme(text = element_text(size = 16))
```

**Automatically handles gene positioning and labeling**
```{=html}
<!-- slide 67 -->
```

## Exercise 1: Your First Volcano Plot - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-7|9-21|23-32"

# Create data
set.seed(789)
my_deg <- data.frame(
  gene = paste0("Gene_", 1:1000),
  log2FC = rnorm(1000, mean = 0, sd = 2),
  pvalue = rbeta(1000, 0.1, 1)
) |>
  mutate(
    padj = p.adjust(pvalue, method = "BH"),
    log10p = -log10(padj),
    status = case_when(
      log2FC > 1.5 & padj < 0.01 ~ "Upregulated",
      log2FC < -1.5 & padj < 0.01 ~ "Downregulated",
      TRUE ~ "Not significant"
    )
  )

# Create plot
ggplot(my_deg, aes(x = log2FC, y = log10p, color = status)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = c("Upregulated" = "#d1422f",
                                 "Downregulated" = "#1a5b5b",
                                 "Not significant" = "gray70")) +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed") +
  labs(title = "My First Volcano Plot",
       x = "log2 Fold Change", y = "-log10(Adjusted P-value)")

# Count genes
table(my_deg$status)
```
```{=html}
<!-- slide 68 -->
```

## Exercise 1: Your First Volcano Plot - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

set.seed(789)
my_deg <- data.frame(
  gene = paste0("Gene_", 1:1000),
  log2FC = rnorm(1000, mean = 0, sd = 2),
  pvalue = rbeta(1000, 0.1, 1)
) |>
  mutate(
    padj = p.adjust(pvalue, method = "BH"),
    log10p = -log10(padj),
    status = case_when(
      log2FC > 1.5 & padj < 0.01 ~ "Upregulated",
      log2FC < -1.5 & padj < 0.01 ~ "Downregulated",
      TRUE ~ "Not significant"
    )
  )

ggplot(my_deg, aes(x = log2FC, y = log10p, color = status)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = c("Upregulated" = "#d1422f",
                                 "Downregulated" = "#1a5b5b",
                                 "Not significant" = "gray70")) +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed") +
  labs(title = "My First Volcano Plot",
       x = "log2 Fold Change", 
       y = "-log10(Adjusted P-value)",
       color = NULL) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```
```{r}
#| echo: false
table(my_deg$status)
```
```{=html}
<!-- slide 69 -->
```

## Exercise 2: Multi-Level Classification - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-10|12-18|20-28"

# Multi-level classification
my_deg <- my_deg |>
  mutate(
    detailed = case_when(
      log2FC > 2 & padj < 0.001 ~ "Highly up",
      log2FC > 1 & padj < 0.05 ~ "Moderately up",
      log2FC < -2 & padj < 0.001 ~ "Highly down",
      log2FC < -1 & padj < 0.05 ~ "Moderately down",
      TRUE ~ "Not significant"
    )
  )

# Custom colors
my_colors <- c("Highly up" = "#d73027",
               "Moderately up" = "#fc8d59",
               "Not significant" = "#cccccc",
               "Moderately down" = "#91bfdb",
               "Highly down" = "#4575b4")

# Plot
ggplot(my_deg, aes(x = log2FC, y = log10p, color = detailed)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = my_colors) +
  geom_hline(yintercept = c(-log10(0.05), -log10(0.001)), 
             linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = c(-2, -1, 1, 2), linetype = "dashed", alpha = 0.5) +
  labs(title = "Multi-Level Classification",
       x = "log2 Fold Change", y = "-log10(Adjusted P-value)")
```
```{=html}
<!-- slide 70 -->
```

## Exercise 2: Multi-Level Classification - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

my_deg <- my_deg |>
  mutate(
    detailed = case_when(
      log2FC > 2 & padj < 0.001 ~ "Highly up",
      log2FC > 1 & padj < 0.05 ~ "Moderately up",
      log2FC < -2 & padj < 0.001 ~ "Highly down",
      log2FC < -1 & padj < 0.05 ~ "Moderately down",
      TRUE ~ "Not significant"
    )
  )

my_colors <- c("Highly up" = "#d73027",
               "Moderately up" = "#fc8d59",
               "Not significant" = "#cccccc",
               "Moderately down" = "#91bfdb",
               "Highly down" = "#4575b4")

ggplot(my_deg, aes(x = log2FC, y = log10p, color = detailed)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = my_colors) +
  geom_hline(yintercept = c(-log10(0.05), -log10(0.001)), 
             linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = c(-2, -1, 1, 2), linetype = "dashed", alpha = 0.5) +
  labs(title = "Multi-Level Classification",
       x = "log2 Fold Change", 
       y = "-log10(Adjusted P-value)",
       color = NULL) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 71 -->
```

## Exercise 3: Faceted Comparisons - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-14|16-21|23-29"

# Create 3 treatment comparisons
set.seed(999)
create_comp <- function(name) {
  data.frame(
    gene = paste0("Gene_", 1:1000),
    log2FC = rnorm(1000, 0, 2),
    pvalue = rbeta(1000, 0.1, 1),
    comparison = name
  ) |>
    mutate(padj = p.adjust(pvalue, "BH"),
           log10p = -log10(padj),
           sig = ifelse(abs(log2FC) > 1 & padj < 0.05, "Sig", "NS"))
}

comp_data <- bind_rows(
  create_comp("Treatment A"),
  create_comp("Treatment B"),
  create_comp("Treatment C")
)

# Faceted plot
ggplot(comp_data, aes(x = log2FC, y = log10p, color = sig)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("Sig" = "#d1422f", "NS" = "gray80")) +
  facet_wrap(~comparison, ncol = 3) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  labs(title = "Multiple Treatment Comparisons",
       x = "log2 Fold Change", y = "-log10(Adjusted P-value)") +
  theme_minimal()
```
```{=html}
<!-- slide 72 -->
```

## Exercise 3: Faceted Comparisons - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

set.seed(999)
create_comp <- function(name) {
  data.frame(
    gene = paste0("Gene_", 1:1000),
    log2FC = rnorm(1000, 0, 2),
    pvalue = rbeta(1000, 0.1, 1),
    comparison = name
  ) |>
    mutate(padj = p.adjust(pvalue, "BH"),
           log10p = -log10(padj),
           sig = ifelse(abs(log2FC) > 1 & padj < 0.05, "Sig", "NS"))
}

comp_data <- bind_rows(
  create_comp("Treatment A"),
  create_comp("Treatment B"),
  create_comp("Treatment C")
)

ggplot(comp_data, aes(x = log2FC, y = log10p, color = sig)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("Sig" = "#d1422f", "NS" = "gray80")) +
  facet_wrap(~comparison, ncol = 3) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  labs(title = "Multiple Treatment Comparisons",
       x = "log2 Fold Change", 
       y = "-log10(Adjusted P-value)",
       color = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```
```{=html}
<!-- slide 73 -->
```





```{=html}
<!-- 


██████╗ ███████╗ █████╗ ██╗  ██╗███████╗
██╔══██╗██╔════╝██╔══██╗██║ ██╔╝██╔════╝
██████╔╝█████╗  ███████║█████╔╝ ███████╗
██╔═══╝ ██╔══╝  ██╔══██║██╔═██╗ ╚════██║
██║     ███████╗██║  ██║██║  ██╗███████║
╚═╝     ╚══════╝╚═╝  ╚═╝╚═╝  ╚═╝╚══════╝
                                        


-->
```

```{=html}
<!-- slide 74 -->
```

# Peak Visualization {background-color="#28A87D"}
```{=html}
<!-- slide 75 -->
```

## Peak Visualization: Introduction

**What are peaks?**

- Regions of enriched signal in genomic assays
- ChIP-seq: Protein-DNA binding sites
- ATAC-seq: Open chromatin regions
- Cut&Run/Cut&Tag: Targeted chromatin profiling

::: fragment
**Why visualize peaks?**

- Understand genomic context (genes, enhancers, promoters)
- Compare conditions or samples
- Validate computational findings
- Generate publication figures
:::
```{=html}
<!-- slide 76 -->
```

## Types of Peak Visualizations

**Two main approaches:**

**Track plots:**
- Show signal across specific genomic regions
- Display multiple samples/conditions
- Include gene annotations
- Zoom to loci of interest

::: fragment
**Genome-wide views:**
- Karyoplots/ideograms
- Manhattan-style plots
- Whole chromosome views
- Overview of peak distribution across genome
:::
```{=html}
<!-- slide 77 -->
```

## Basic Peak Track: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-5|7-11|13-20"

# Simulate genomic signal data
set.seed(123)
genomic_pos <- 1:1000
signal <- abs(rnorm(1000, mean = 2, sd = 3))
signal[450:550] <- signal[450:550] + rnorm(101, mean = 10, sd = 2)

# Create data frame
track_data <- data.frame(
  position = genomic_pos,
  signal = signal
)
head(track_data)

# Plot peak track
ggplot(track_data, aes(x = position, y = signal)) +
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "darkblue", linewidth = 0.5) +
  annotate("rect", xmin = 450, xmax = 550, ymin = 0, ymax = Inf,
           fill = "#d1422f", alpha = 0.1) +
  labs(title = "Example Peak Track",
       x = "Genomic Position (kb)", y = "Signal Intensity")+
  custom_theme()
```
```{=html}
<!-- slide 78 -->
```

## Basic Peak Track: Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center

set.seed(123)
genomic_pos <- 1:1000
signal <- abs(rnorm(1000, mean = 2, sd = 3))
signal[450:550] <- signal[450:550] + rnorm(101, mean = 10, sd = 2)

track_data <- data.frame(
  position = genomic_pos,
  signal = signal
)

ggplot(track_data, aes(x = position, y = signal)) +
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "darkblue", linewidth = 0.5) +
  annotate("rect", xmin = 450, xmax = 550, ymin = 0, ymax = Inf,
           fill = "#d1422f", alpha = 0.1) +
  annotate("text", x = 500, y = 18, label = "Peak", color = "#d1422f", size = 5) +
  labs(
    title = "Example Peak Track",
    x = "Genomic Position (kb)",
    y = "Signal Intensity"
  ) 
```

```{=html}
<!-- slide 79 -->
```
## What is a karyotype plot?

A **karyotype plot** visualizes data along **real chromosome structures**.

- Chromosomes are drawn to scale
- Genomic coordinates are preserved
- Signals are mapped onto physical chromosomes

::: fragment
**Typical questions it answers**

- Where do signals cluster in the genome?
- Are specific chromosomes enriched?
- Do signals accumulate near chromosome ends?
:::

```{=html}
<!-- slide 80 -->
```
## Method 1: karyoploteR - Introduction

**karyoploteR: Genome-wide visualization**

- Publication-quality ideograms
- Whole genome or chromosome-specific views
- Multiple data tracks on karyotypes
- Customizable layouts (7 different plot types)

::: fragment
**Best for:**

- Showing peak distribution across chromosomes
- Comparing datasets genome-wide
- Creating overview figures
- CNV visualization
:::
```{=html}
<!-- slide 81 -->
```
## Which organisms can be plotted?

karyoploteR works best with organisms that have:

- A reference genome
- Known chromosome lengths

Common examples:

- Human (`hg19`, `hg38`)
- Mouse (`mm10`)
- Zebrafish
- Drosophila
- Arabidopsis
- Yeast

::: fragment
💡 If an organism has a UCSC or Ensembl genome, it usually works.
:::

```{=html}
<!-- slide 82 -->
```

## Using your own organism

Any organism can be plotted if you provide:

- Chromosome names
- Chromosome lengths

karyoploteR only needs chromosome boundaries —  
not gene annotations.
```{r}
custom.genome <- data.frame(
  chr = "chr1",
  start = 1,
  end = 4600000
)

plotKaryotype(genome = custom.genome)
```


## karyoploteR: Basic Karyoplot - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-2|4-11|13-15"

# Load packages
library(karyoploteR)
library(GenomicRanges)

# Create example peak data
set.seed(123)
peaks <- GRanges(
  seqnames = paste0("chr", sample(1:22, 100, replace = TRUE)),
  ranges = IRanges(start = sample(1:200000000, 100),
                   width = sample(200:2000, 100)),
  score = runif(100, 0, 10)
)

# Create karyoplot and add peaks
kp <- plotKaryotype(genome = "hg38", chromosomes = paste0("chr", 1:22))
kpPlotRegions(kp, data = peaks, col = "steelblue")
```


```{=html}
<!-- slide 81 -->
```

## karyoploteR: Basic Karyoplot - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(karyoploteR)
library(GenomicRanges)

set.seed(123)
peaks <- GRanges(
  seqnames = paste0("chr", sample(1:22, 100, replace = TRUE)),
  ranges = IRanges(
    start = sample(1:200000000, 100),
    width = sample(200:2000, 100)
  ),
  score = runif(100, 0, 10)
)

kp <- plotKaryotype(genome = "hg38", chromosomes = paste0("chr", 1:22))
kpPlotRegions(kp, data = peaks, col = "steelblue")
```

```{=html}
<!-- slide 82 -->
```

## karyoploteR: Multi-Track - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-2|4-7|9-12|14-17"

# Create karyoplot
kp <- plotKaryotype(genome = "hg38", chromosomes = paste0("chr", 1:5))

# Track 1: Sample A peaks (top)
kpPlotRegions(kp, data = peaks, 
              r0 = 0, r1 = 0.3,
              col = "#d1422f")

# Track 2: Sample B peaks (middle)
kpPlotRegions(kp, data = peaks, 
              r0 = 0.35, r1 = 0.65,
              col = "#1a5b5b")

# Track 3: Density (bottom)
kpPlotDensity(kp, data = peaks, 
              r0 = 0.7, r1 = 1,
              col = "#f4ab5c")
```

**r0/r1:** Vertical position (0 = bottom, 1 = top)
```{=html}
<!-- slide 83 -->
```

## karyoploteR: Multi-Track - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(karyoploteR)
library(GenomicRanges)

set.seed(123)
peaks <- GRanges(
  seqnames = paste0("chr", sample(1:5, 100, replace = TRUE)),
  ranges = IRanges(
    start = sample(1:200000000, 100),
    width = sample(200:2000, 100)
  ),
  score = runif(100, 0, 10)
)

kp <- plotKaryotype(genome = "hg38", chromosomes = paste0("chr", 1:5))
kpPlotRegions(kp, data = peaks, r0 = 0, r1 = 0.3, col = "#d1422f")
kpPlotRegions(kp, data = peaks, r0 = 0.35, r1 = 0.65, col = "#1a5b5b")
kpPlotDensity(kp, data = peaks, r0 = 0.7, r1 = 1, col = "#f4ab5c")
```
```{=html}
<!-- slide 84 -->
```

## karyoploteR: Density Plot - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-2|4-8"

# Create karyoplot
kp <- plotKaryotype(genome = "hg38", chromosomes = "chr1")

# Add peak density with window
kpPlotDensity(kp, 
              data = peaks, 
              window.size = 1e7,  # 10Mb windows
              col = "steelblue")
```

**window.size:** Controls smoothing of density plot
```{=html}
<!-- slide 85 -->
```

## karyoploteR: Density Plot - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(karyoploteR)
library(GenomicRanges)

set.seed(456)
peaks_chr1 <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(
    start = sample(1:200000000, 200),
    width = sample(200:2000, 200)
  )
)

kp <- plotKaryotype(genome = "hg38", chromosomes = "chr1")
kpPlotDensity(kp, data = peaks_chr1, window.size = 1e7, col = "steelblue")
```
```{=html}
<!-- slide 86 -->
```

## karyoploteR: Coverage Plot - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-9|11-15"

# Create coverage data
set.seed(789)
coverage <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(start = seq(1e6, 5e6, by = 1000),
                   width = 1000),
  score = abs(rnorm(4001, mean = 50, sd = 20))
)

# Plot coverage
kp <- plotKaryotype(genome = "hg38", chromosomes = "chr1")
kpPlotCoverage(kp, 
               data = coverage,
               col = "forestgreen",
               r0 = 0, r1 = 0.5)
```
```{=html}
<!-- slide 87 -->
```

## karyoploteR: Coverage Plot - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(karyoploteR)
library(GenomicRanges)

set.seed(789)
coverage <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(start = seq(1e6, 5e6, by = 1000),
                   width = 1000),
  score = abs(rnorm(4001, mean = 50, sd = 20))
)

kp <- plotKaryotype(genome = "hg38", chromosomes = "chr1", plot.type = 4)
kpPlotCoverage(kp, data = coverage, col = "forestgreen", r0 = 0, r1 = 0.5)
```
```{=html}
<!-- slide 88 -->
```

## karyoploteR: Chromosome Zoom - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-5|7-10|12-13"

# Focus on specific region
kp <- plotKaryotype(
  genome = "hg38", 
  chromosomes = "chr1",
  plot.type = 4
)

# Add multiple data types
kpPlotRegions(kp, data = peaks, col = "#d1422f", r0 = 0, r1 = 0.45)
kpPlotDensity(kp, data = peaks, col = "#1a5b5b", r0 = 0.55, r1 = 1)

# Add horizontal line at specific position
kpAbline(kp, h = 0.5, col = "gray", lty = 2)
```

**plot.type 4:** Horizontal layout optimized for single chromosome
```{=html}
<!-- slide 89 -->
```

## karyoploteR: Chromosome Zoom - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(karyoploteR)

set.seed(456)
peaks_chr1 <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(
    start = sample(1:200000000, 200),
    width = sample(200:2000, 200)
  )
)

kp <- plotKaryotype(genome = "hg38", chromosomes = "chr1", plot.type = 4)
kpPlotRegions(kp, data = peaks_chr1, col = "#d1422f", r0 = 0, r1 = 0.45)
kpPlotDensity(kp, data = peaks_chr1, col = "#1a5b5b", r0 = 0.55, r1 = 1, window.size = 1e7)
```
```{=html}
<!-- slide 90 -->
```

## Method 2: Gviz - Introduction

**Gviz: Detailed locus-level visualization**

- Multiple track types (data, annotation, genes, sequence)
- Fine control over every visual element
- Publication-ready track plots
- Integration with Bioconductor annotations

::: fragment
**Best for:**

- Detailed views of specific loci
- ChIP-seq with gene models
- Multi-sample comparisons
- Publication figures
:::
```{=html}
<!-- slide 91 -->
```

## Gviz: Basic Tracks - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-2|4-10|12-18|20-26"

# Load packages
library(Gviz)
library(GenomicRanges)

# Create example ChIP-seq peaks
chipseq_peaks <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(start = c(1000000, 1005000, 1010000),
                   end = c(1002000, 1007000, 1012000)),
  score = c(25, 45, 35)
)

# Create tracks
genome_axis <- GenomeAxisTrack()
ideogram <- IdeogramTrack(genome = "hg38", chromosome = "chr1")
peaks_track <- AnnotationTrack(
  chipseq_peaks, name = "Peaks", fill = "steelblue"
)

# Plot
plotTracks(
  list(ideogram, genome_axis, peaks_track),
  from = 1000000, 
  to = 1015000,
  chromosome = "chr1"
)
```
```{=html}
<!-- slide 92 -->
```

## Gviz: Basic Tracks - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(Gviz)
library(GenomicRanges)

chipseq_peaks <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(
    start = c(1000000, 1005000, 1010000),
    end = c(1002000, 1007000, 1012000)
  ),
  score = c(25, 45, 35)
)

genome_axis <- GenomeAxisTrack()
ideogram <- IdeogramTrack(genome = "hg38", chromosome = "chr1")
peaks_track <- AnnotationTrack(
  chipseq_peaks, 
  name = "Peaks",
  fill = "steelblue"
)

plotTracks(
  list(ideogram, genome_axis, peaks_track),
  from = 1000000, to = 1015000,
  chromosome = "chr1"
)
```
```{=html}
<!-- slide 93 -->
```

## Gviz: Data Tracks - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-9|11-17|19-25"

# Create coverage data for sample
set.seed(123)
coverage_data <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(start = seq(1e6, 1.5e6, by = 100),
                   width = 100),
  score = abs(rnorm(5001, mean = 50, sd = 20))
)

# Create DataTrack
coverage_track <- DataTrack(
  coverage_data,
  name = "ChIP Signal",
  type = "histogram",
  fill = "darkgreen",
  col = "darkgreen"
)

# Plot with axis
plotTracks(
  list(ideogram, genome_axis, coverage_track),
  from = 1e6,
  to = 1.5e6,
  chromosome = "chr1"
)
```
```{=html}
<!-- slide 94 -->
```

## Gviz: Data Tracks - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(Gviz)

set.seed(123)
coverage_data <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(start = seq(1e6, 1.5e6, by = 100),
                   width = 100),
  score = abs(rnorm(5001, mean = 50, sd = 20))
)

genome_axis <- GenomeAxisTrack()
ideogram <- IdeogramTrack(genome = "hg38", chromosome = "chr1")

coverage_track <- DataTrack(
  coverage_data,
  name = "ChIP Signal",
  type = "histogram",
  fill = "darkgreen",
  col = "darkgreen"
)

plotTracks(
  list(ideogram, genome_axis, coverage_track),
  from = 1e6, to = 1.5e6,
  chromosome = "chr1"
)
```
```{=html}
<!-- slide 95 -->
```

## Gviz: Multi-Sample Comparison - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-7|9-15|17-25"

# Sample 1: Control
sample1_track <- DataTrack(
  coverage_data,
  name = "Control",
  type = "histogram",
  fill = "gray60"
)

# Sample 2: Treatment (higher signal)
sample2_track <- DataTrack(
  coverage_data,
  name = "Treatment",
  type = "histogram",
  fill = "firebrick"
)

# Plot both samples
plotTracks(
  list(ideogram, genome_axis,
       sample1_track,
       sample2_track),
  from = 1e6,
  to = 1.5e6,
  chromosome = "chr1"
)
```
```{=html}
<!-- slide 96 -->
```

## Gviz: Multi-Sample Comparison - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(Gviz)

set.seed(123)
coverage_data <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(start = seq(1e6, 1.5e6, by = 100),
                   width = 100),
  score = abs(rnorm(5001, mean = 50, sd = 20))
)

genome_axis <- GenomeAxisTrack()
ideogram <- IdeogramTrack(genome = "hg38", chromosome = "chr1")

sample1_track <- DataTrack(coverage_data, name = "Control",
                          type = "histogram", fill = "gray60")
sample2_track <- DataTrack(coverage_data, name = "Treatment",
                          type = "histogram", fill = "firebrick")

plotTracks(
  list(ideogram, genome_axis, sample1_track, sample2_track),
  from = 1e6, to = 1.5e6, chromosome = "chr1"
)
```
```{=html}
<!-- slide 97 -->
```

## Gviz: Different Plot Types - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-5|7-11|13-17|19-23"

# Type: polygon (filled area)
track1 <- DataTrack(coverage_data, name = "Polygon",
                   type = "polygon", fill = "lightblue")

# Type: line
track2 <- DataTrack(coverage_data, name = "Line",
                   type = "l", col = "darkblue", lwd = 2)

# Type: points
track3 <- DataTrack(coverage_data, name = "Points",
                   type = "p", col = "#d1422f")

# Type: smooth (smoothed line)
track4 <- DataTrack(coverage_data, name = "Smooth",
                   type = "smooth", col = "purple")

# Plot all
plotTracks(list(ideogram, genome_axis, track1, track2, track3, track4),
           from = 1e6, to = 1.5e6, chromosome = "chr1")
```

**Available types:** histogram, polygon, line, points, smooth, mountain, heatmap
```{=html}
<!-- slide 98 -->
```

## Gviz: Different Plot Types - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 8
#| fig-align: center
#| message: false
#| warning: false

library(Gviz)

set.seed(456)
coverage_data <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(start = seq(1e6, 1.2e6, by = 100),
                   width = 100),
  score = abs(rnorm(2001, mean = 50, sd = 20))
)

genome_axis <- GenomeAxisTrack()
ideogram <- IdeogramTrack(genome = "hg38", chromosome = "chr1")

track1 <- DataTrack(coverage_data, name = "Polygon", type = "polygon", fill = "lightblue")
track2 <- DataTrack(coverage_data, name = "Line", type = "l", col = "darkblue", lwd = 2)

plotTracks(list(ideogram, genome_axis, track1, track2),
           from = 1e6, to = 1.2e6, chromosome = "chr1")
```
```{=html}
<!-- slide 99 -->
```

## Gviz: Overlay Tracks - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-10|12-18|20-27"

# Create overlapping data
set.seed(999)
data1 <- GRanges("chr1", IRanges(seq(1e6, 1.5e6, 100), width = 100),
                score = abs(rnorm(5001, 50, 20)))
data2 <- GRanges("chr1", IRanges(seq(1e6, 1.5e6, 100), width = 100),
                score = abs(rnorm(5001, 40, 15)))

# Create overlay track
overlay_track <- OverlayTrack(
  trackList = list(
    DataTrack(data1, type = "l", col = "#d1422f", name = "Sample1"),
    DataTrack(data2, type = "l", col = "#1a5b5b", name = "Sample2")
  )
)

# Plot
plotTracks(
  list(ideogram, genome_axis, overlay_track),
  from = 1e6,
  to = 1.5e6,
  chromosome = "chr1"
)
```

**OverlayTrack:** Combine multiple data tracks in same space
```{=html}
<!-- slide 100 -->
```

## Gviz: Overlay Tracks - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(Gviz)

set.seed(999)
data1 <- GRanges("chr1", IRanges(seq(1e6, 1.5e6, 500), width = 100),
                score = abs(rnorm(1001, 50, 20)))
data2 <- GRanges("chr1", IRanges(seq(1e6, 1.5e6, 500), width = 100),
                score = abs(rnorm(1001, 40, 15)))

genome_axis <- GenomeAxisTrack()
ideogram <- IdeogramTrack(genome = "hg38", chromosome = "chr1")

overlay_track <- OverlayTrack(
  trackList = list(
    DataTrack(data1, type = "l", col = "#d1422f", name = "Sample1"),
    DataTrack(data2, type = "l", col = "#1a5b5b", name = "Sample2")
  )
)

plotTracks(
  list(ideogram, genome_axis, overlay_track),
  from = 1e6, to = 1.5e6, chromosome = "chr1"
)
```
```{=html}
<!-- slide 101 -->
```

## Gviz Track Types Summary

**Available track types:**

- `DataTrack`: Signal/coverage (7+ plot types)
- `AnnotationTrack`: Peaks, regions, features
- `GeneRegionTrack`: Gene models with exons
- `SequenceTrack`: DNA sequence
- `OverlayTrack`: Combine multiple tracks
- `AlignmentsTrack`: BAM file alignments

::: fragment
**Key parameters:**

- `type`: Plot style (histogram, line, polygon, smooth, etc.)
- `fill`/`col`: Colors
- `from`/`to`: Genomic coordinates to display
- Track-specific heights and customizations
:::
```{=html}
<!-- slide 102 -->
```

## Method Comparison
```{r}
#| echo: false

library(gt)

comparison <- tibble(
  Package = c("karyoploteR", "Gviz", "Signac"),
  `Best For` = c(
    "Genome-wide views, chromosome patterns",
    "Detailed locus views, multi-sample",
    "Single-cell ATAC-seq, cell types"
  ),
  Complexity = c("Medium", "High", "High"),
  Flexibility = c("High", "Very High", "Medium"),
  `Learning Curve` = c("Moderate", "Steep", "Moderate")
)

comparison |>
  gt() |>
  cols_align(align = "left") |>
  tab_style(
    style = cell_text(weight = "bold", size = px(14)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(size = px(12)),
    locations = cells_body()
  ) |>
  tab_options(
    table.width = pct(95)
  )
```



```{=html}

<!--
   _____                                 _                     
  / ____|                               (_)                    
 | |     ___  _ __ ___  _ __   __ _ _ __ _ ___  ___  _ __  ___ 
 | |    / _ \| '_ ` _ \| '_ \ / _` | '__| / __|/ _ \| '_ \/ __|
 | |___| (_) | | | | | | |_) | (_| | |  | \__ \ (_) | | | \__ \
  \_____\___/|_| |_| |_| .__/ \__,_|_|  |_|___/\___/|_| |_|___/
                       | |                                     
                       |_|            
-->

```

# Advanced Visualizations {background-color="#28A87D"}
```{=html}
<!-- slide 105 -->
```

## Heatmaps: Introduction

**What are heatmaps?**

- Matrix visualization using color intensity
- Rows: genes, samples, or features
- Columns: samples, conditions, or measurements
- Color: expression level or correlation

::: fragment
**Common uses in genomics:**

- Gene expression patterns across samples
- Sample clustering and relationships
- Correlation matrices
- Methylation patterns
:::
```{=html}
<!-- slide 106 -->
```

## Why Use Heatmaps?

**Advantages:**

- Visualize thousands of genes simultaneously
- Reveal patterns and clusters
- Show hierarchical relationships
- Compare multiple conditions at once

::: fragment
**Key components:**

- Dendrograms (clustering trees)
- Color scales (sequential or diverging)
- Row/column annotations
- Z-score normalization
:::
```{=html}
<!-- slide 107 -->
```

## Basic Heatmap: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-9|11-17|19-23"

# Simulate gene expression matrix
set.seed(123)
n_genes <- 50
n_samples <- 10

expr_matrix <- matrix(
  rnorm(n_genes * n_samples, mean = 50, sd = 20),
  nrow = n_genes,
  dimnames = list(paste0("Gene_", 1:n_genes),
                  paste0("Sample_", 1:n_samples))
)

# Add tissue-specific patterns
expr_matrix[1:20, 1:5] <- expr_matrix[1:20, 1:5] + 40  # Brain
expr_matrix[21:40, 6:10] <- expr_matrix[21:40, 6:10] + 40  # Liver

# Create heatmap
library(pheatmap)
pheatmap(expr_matrix,
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         main = "Gene Expression Heatmap")
```
```{=html}
<!-- slide 108 -->
```

## Basic Heatmap: Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center
#| message: false
#| warning: false

library(pheatmap)

set.seed(123)
n_genes <- 50
n_samples <- 10

expr_matrix <- matrix(
  rnorm(n_genes * n_samples, mean = 50, sd = 20),
  nrow = n_genes,
  dimnames = list(paste0("Gene_", 1:n_genes),
                  paste0("Sample_", 1:n_samples))
)

expr_matrix[1:20, 1:5] <- expr_matrix[1:20, 1:5] + 40
expr_matrix[21:40, 6:10] <- expr_matrix[21:40, 6:10] + 40

pheatmap(expr_matrix,
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         main = "Gene Expression Heatmap",
         fontsize = 10)
```
```{=html}
<!-- slide 109 -->
```

## Annotated Heatmap: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-5|7-11|13-20"

# Create sample annotations
sample_anno <- data.frame(
  Tissue = rep(c("Brain", "Liver"), each = 5),
  Treatment = rep(c("Control", "Treated"), times = 5),
  row.names = colnames(expr_matrix)
)

# Define annotation colors
anno_colors <- list(
  Tissue = c(Brain = "#E64B35", Liver = "#4DBBD5"),
  Treatment = c(Control = "gray70", Treated = "gold")
)

# Create annotated heatmap
pheatmap(expr_matrix,
         scale = "row",
         annotation_col = sample_anno,
         annotation_colors = anno_colors,
         show_rownames = FALSE,
         main = "Annotated Gene Expression")
```
```{=html}
<!-- slide 110 -->
```

## Annotated Heatmap: Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center
#| message: false
#| warning: false

sample_anno <- data.frame(
  Tissue = rep(c("Brain", "Liver"), each = 5),
  Treatment = rep(c("Control", "Treated"), times = 5),
  row.names = colnames(expr_matrix)
)

anno_colors <- list(
  Tissue = c(Brain = "#E64B35", Liver = "#4DBBD5"),
  Treatment = c(Control = "gray70", Treated = "gold")
)

pheatmap(expr_matrix,
         scale = "row",
         annotation_col = sample_anno,
         annotation_colors = anno_colors,
         show_rownames = FALSE,
         main = "Annotated Gene Expression",
         fontsize = 10)
```
```{=html}
<!-- slide 111 -->
```

## ComplexHeatmap: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-2|4-11|13-19"

# Load ComplexHeatmap
library(ComplexHeatmap)
library(circlize)

# Create color function
col_fun <- colorRamp2(
  c(-2, 0, 2), 
  c("#1a5b5b", "white", "#d1422f")
)

# Create complex heatmap
Heatmap(scale(expr_matrix),
        name = "Z-score",
        col = col_fun,
        top_annotation = HeatmapAnnotation(
          Tissue = sample_anno$Tissue,
          col = list(Tissue = c(Brain = "#E64B35", Liver = "#4DBBD5"))
        ),
        show_row_names = FALSE,
        column_title = "Complex Heatmap with Annotations")
```
```{=html}
<!-- slide 112 -->
```

## ComplexHeatmap: Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center
#| message: false
#| warning: false

library(ComplexHeatmap)
library(circlize)

col_fun <- colorRamp2(
  c(-2, 0, 2), 
  c("#1a5b5b", "white", "#d1422f")
)

Heatmap(scale(expr_matrix),
        name = "Z-score",
        col = col_fun,
        top_annotation = HeatmapAnnotation(
          Tissue = sample_anno$Tissue,
          col = list(Tissue = c(Brain = "#E64B35", Liver = "#4DBBD5"))
        ),
        show_row_names = FALSE,
        column_title = "Complex Heatmap with Annotations")
```
```{=html}
<!-- slide 113 -->
```

## Venn Diagrams: Introduction

**What are Venn diagrams?**

- Show overlaps between sets
- Compare gene lists from different experiments
- Visualize shared and unique elements
- Maximum 3-4 sets for clarity

::: fragment
**Common genomics applications:**

- DEGs across multiple comparisons
- Overlapping peaks from ChIP-seq
- Common genes in pathways
- Shared variants between samples
:::
```{=html}
<!-- slide 114 -->
```

## Creating Gene Sets: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-10|12-16|18-20"

# Simulate differentially expressed genes
set.seed(456)

# Create gene universe
all_genes <- paste0("Gene_", 1:1000)

# Treatment A DEGs (200 genes)
treatment_A <- sample(all_genes, 200)

# Treatment B DEGs (250 genes, 100 overlap with A)
treatment_B <- c(
  sample(treatment_A, 100),  # 100 overlap
  sample(setdiff(all_genes, treatment_A), 150)  # 150 unique
)

# Treatment C DEGs (180 genes)
treatment_C <- sample(all_genes, 180)

# Create list
gene_lists <- list(
  Treatment_A = treatment_A,
  Treatment_B = treatment_B,
  Treatment_C = treatment_C
)
```
```{=html}
<!-- slide 115 -->
```

## Venn Diagram: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-2|4-12"

# Load VennDiagram package
library(VennDiagram)

# Create Venn diagram
venn.plot <- venn.diagram(
  x = gene_lists,
  category.names = c("Treatment A", "Treatment B", "Treatment C"),
  filename = NULL,
  fill = c("#E64B35", "#4DBBD5", "#00A087"),
  alpha = 0.5,
  cex = 1.5,
  cat.cex = 1.2
)

# Display
grid.draw(venn.plot)
```
```{=html}
<!-- slide 116 -->
```

## Venn Diagram: Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center
#| message: false
#| warning: false

library(VennDiagram)
library(grid)

set.seed(456)
all_genes <- paste0("Gene_", 1:1000)
treatment_A <- sample(all_genes, 200)
treatment_B <- c(sample(treatment_A, 100), 
                 sample(setdiff(all_genes, treatment_A), 150))
treatment_C <- sample(all_genes, 180)

gene_lists <- list(
  Treatment_A = treatment_A,
  Treatment_B = treatment_B,
  Treatment_C = treatment_C
)

venn.plot <- venn.diagram(
  x = gene_lists,
  category.names = c("Treatment A", "Treatment B", "Treatment C"),
  filename = NULL,
  fill = c("#E64B35", "#4DBBD5", "#00A087"),
  alpha = 0.5,
  cex = 1.5,
  cat.cex = 1.2,
  cat.pos = c(-20, 20, 180)
)

grid.newpage()
grid.draw(venn.plot)
```
```{=html}
<!-- slide 117 -->
```

## UpSet Plots: Introduction

**Why UpSet plots?**

- Better than Venn diagrams for >3 sets
- Show exact intersection sizes
- Clear visualization of complex overlaps
- Scalable to many sets

::: fragment
**When to use:**

- Comparing 4+ gene lists
- Multiple tissue comparisons
- Multi-condition experiments
- Complex set relationships
:::
```{=html}
<!-- slide 118 -->
```

## Creating Multi-Set Data: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-10|12-18"

# Simulate 5 different conditions
set.seed(789)
all_genes <- paste0("Gene_", 1:1000)

gene_sets <- list(
  Brain = sample(all_genes, 250),
  Liver = sample(all_genes, 200),
  Heart = sample(all_genes, 180),
  Kidney = sample(all_genes, 220),
  Lung = sample(all_genes, 190)
)

# Calculate overlaps
library(UpSetR)

# Create UpSet plot
upset(fromList(gene_sets),
      nsets = 5,
      order.by = "freq")
```
```{=html}
<!-- slide 119 -->
```

## UpSet Plot: Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center
#| message: false
#| warning: false

library(UpSetR)

set.seed(789)
all_genes <- paste0("Gene_", 1:1000)

gene_sets <- list(
  Brain = sample(all_genes, 250),
  Liver = sample(all_genes, 200),
  Heart = sample(all_genes, 180),
  Kidney = sample(all_genes, 220),
  Lung = sample(all_genes, 190)
)

upset(fromList(gene_sets),
      nsets = 5,
      order.by = "freq",
      text.scale = 1.5)
```
```{=html}
<!-- slide 120 -->
```

## UpSet Plot with Queries: Code
```{r}
#| eval: false
#| code-line-numbers: "|1-9|11-19"

# Highlight specific intersections
upset(fromList(gene_sets),
      nsets = 5,
      order.by = "freq",
      queries = list(
        list(query = intersects,
             params = list("Brain", "Liver"),
             color = "#d1422f", active = TRUE)
      ))

# Show only specific intersections
upset(fromList(gene_sets),
      nsets = 5,
      nintersects = 20,  # Show top 20 intersections
      mb.ratio = c(0.6, 0.4),  # Adjust matrix/bar ratio
      order.by = "freq",
      text.scale = c(1.5, 1.5, 1.3, 1.3, 1.5, 1.2),
      point.size = 3.5,
      line.size = 1)
```
```{=html}
<!-- slide 121 -->
```

## UpSet Plot with Queries: Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center
#| message: false
#| warning: false

library(UpSetR)

upset(fromList(gene_sets),
      nsets = 5,
      nintersects = 20,
      mb.ratio = c(0.6, 0.4),
      order.by = "freq",
      text.scale = c(1.5, 1.5, 1.3, 1.3, 1.5, 1.2),
      point.size = 3.5,
      line.size = 1)
```
```{=html}
<!-- slide 122 -->
```

## Exercise 1: Custom Heatmap - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-10|12-18|20-27"

# Create cell type-specific expression
set.seed(111)
n_genes <- 100
n_cells <- 30

expr_data <- matrix(
  rnorm(n_genes * n_cells, mean = 5, sd = 2),
  nrow = n_genes,
  dimnames = list(paste0("Gene_", 1:n_genes),
                  paste0("Cell_", 1:n_cells))
)

# Add cell type patterns
expr_data[1:30, 1:10] <- expr_data[1:30, 1:10] + 5   # T cells
expr_data[31:60, 11:20] <- expr_data[31:60, 11:20] + 5  # B cells
expr_data[61:100, 21:30] <- expr_data[61:100, 21:30] + 5  # Monocytes

# Create annotations
cell_anno <- data.frame(
  CellType = c(rep("T cell", 10), rep("B cell", 10), rep("Monocyte", 10)),
  Donor = rep(paste0("Donor", 1:3), length.out = 30),
  row.names = colnames(expr_data)
)

# Plot
library(pheatmap)
pheatmap(expr_data, scale = "row",
         annotation_col = cell_anno,
         show_rownames = FALSE,
         main = "Single-Cell Expression Patterns")
```
```{=html}
<!-- slide 123 -->
```

## Exercise 1: Custom Heatmap - Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center
#| message: false
#| warning: false

library(pheatmap)

set.seed(111)
n_genes <- 100
n_cells <- 30

expr_data <- matrix(
  rnorm(n_genes * n_cells, mean = 5, sd = 2),
  nrow = n_genes,
  dimnames = list(paste0("Gene_", 1:n_genes),
                  paste0("Cell_", 1:n_cells))
)

expr_data[1:30, 1:10] <- expr_data[1:30, 1:10] + 5
expr_data[31:60, 11:20] <- expr_data[31:60, 11:20] + 5
expr_data[61:100, 21:30] <- expr_data[61:100, 21:30] + 5

cell_anno <- data.frame(
  CellType = c(rep("T cell", 10), rep("B cell", 10), rep("Monocyte", 10)),
  Donor = rep(paste0("Donor", 1:3), length.out = 30),
  row.names = colnames(expr_data)
)

pheatmap(expr_data, scale = "row",
         annotation_col = cell_anno,
         show_rownames = FALSE,
         main = "Single-Cell Expression Patterns",
         fontsize = 10)
```
```{=html}
<!-- slide 124 -->
```

## Exercise 2: Correlation Heatmap - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-5|7-12|14-19"

# Calculate sample correlations
cor_matrix <- cor(expr_matrix)

# Create correlation heatmap
library(pheatmap)

# Custom colors for correlation
library(RColorBrewer)
cor_colors <- colorRampPalette(rev(brewer.pal(9, "RdBu")))(100)

pheatmap(cor_matrix,
         color = cor_colors,
         breaks = seq(-1, 1, length.out = 101),
         display_numbers = TRUE,
         number_format = "%.2f",
         fontsize_number = 10,
         main = "Sample Correlation Heatmap")
```
```{=html}
<!-- slide 125 -->
```

## Exercise 2: Correlation Heatmap - Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center
#| message: false
#| warning: false

library(pheatmap)
library(RColorBrewer)

cor_matrix <- cor(expr_matrix)

cor_colors <- colorRampPalette(rev(brewer.pal(9, "RdBu")))(100)

pheatmap(cor_matrix,
         color = cor_colors,
         breaks = seq(-1, 1, length.out = 101),
         display_numbers = TRUE,
         number_format = "%.2f",
         fontsize_number = 10,
         main = "Sample Correlation Heatmap")
```
```{=html}
<!-- slide 126 -->
```

## Exercise 3: Pathway Overlap - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-9|11-17|19-24"

# Simulate pathway analysis results
set.seed(222)
all_genes <- paste0("Gene_", 1:500)

pathways <- list(
  "Cell Cycle" = sample(all_genes, 80),
  "Apoptosis" = sample(all_genes, 70),
  "Immune Response" = sample(all_genes, 90),
  "Metabolism" = sample(all_genes, 100)
)

# Create Venn for 2 pathways
library(VennDiagram)
venn.diagram(
  x = list(CellCycle = pathways[[1]],
           Apoptosis = pathways[[2]]),
  filename = NULL,
  fill = c("lightblue", "pink"),
  alpha = 0.5
)

# For all 4 pathways, use UpSet
library(UpSetR)
upset(fromList(pathways),
      order.by = "freq",
      text.scale = 1.5)
```
```{=html}
<!-- slide 127 -->
```

## Exercise 3: Pathway Overlap - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center
#| message: false
#| warning: false

library(UpSetR)

set.seed(222)
all_genes <- paste0("Gene_", 1:500)

pathways <- list(
  "Cell Cycle" = sample(all_genes, 80),
  "Apoptosis" = sample(all_genes, 70),
  "Immune Response" = sample(all_genes, 90),
  "Metabolism" = sample(all_genes, 100)
)

upset(fromList(pathways),
      order.by = "freq",
      text.scale = 1.5,
      point.size = 3.5,
      line.size = 1)
```
```{=html}
<!-- slide 128 -->
```

## Exercise 4: Multi-Condition Comparison - Code
```{r}
#| eval: false
#| code-line-numbers: "|1-12|14-22"

# Simulate DEGs from multiple comparisons
set.seed(333)
all_genes <- paste0("Gene_", 1:800)

deg_sets <- list(
  "Condition1_vs_Control" = sample(all_genes, 150),
  "Condition2_vs_Control" = sample(all_genes, 180),
  "Condition3_vs_Control" = sample(all_genes, 160),
  "Condition4_vs_Control" = sample(all_genes, 140),
  "Condition5_vs_Control" = sample(all_genes, 170)
)

# Create comprehensive UpSet plot
library(UpSetR)
upset(fromList(deg_sets),
      nsets = 5,
      nintersects = 30,
      order.by = "freq",
      mb.ratio = c(0.6, 0.4),
      text.scale = 1.3,
      point.size = 3,
      main.bar.color = "steelblue")
```
```{=html}
<!-- slide 129 -->
```

## Exercise 4: Multi-Condition Comparison - Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7.5
#| fig-align: center
#| message: false
#| warning: false

library(UpSetR)

set.seed(333)
all_genes <- paste0("Gene_", 1:800)

deg_sets <- list(
  "Cond1_vs_Ctrl" = sample(all_genes, 150),
  "Cond2_vs_Ctrl" = sample(all_genes, 180),
  "Cond3_vs_Ctrl" = sample(all_genes, 160),
  "Cond4_vs_Ctrl" = sample(all_genes, 140),
  "Cond5_vs_Ctrl" = sample(all_genes, 170)
)

upset(fromList(deg_sets),
      nsets = 5,
      nintersects = 30,
      order.by = "freq",
      mb.ratio = c(0.6, 0.4),
      text.scale = 1.3,
      point.size = 3,
      main.bar.color = "steelblue")
```
```{=html}
<!-- slide 130 -->
```

## Summary: When to Use Each

**Heatmaps:**
- Gene expression across samples
- Correlation matrices
- Clustering visualization
- Any matrix data

**Venn Diagrams:**
- 2-3 set comparisons
- Simple overlaps
- Quick visual summary

**UpSet Plots:**
- 4+ set comparisons
- Complex intersections
- Detailed overlap analysis
- Scalable visualization
```{=html}
<!-- slide 131 -->
```


```{=html}
<!-- 
slide 60
-->
```

# Phylogenetic trees {background-color="#28A87D"}
```{=html}
<!-- 
slide 61
-->
```

## Why Visualize Phylogenetic Trees?

**Understanding evolutionary relationships:**

::: fragment
**Key applications:**

- Show species relationships and divergence
- Identify conserved and divergent lineages
- Detect horizontal gene transfer
- Guide taxonomic classification
- Compare gene family evolution
:::

::: fragment
**Visualization challenges:**

- Trees can have hundreds to thousands of tips
- Multiple layout options for different insights
- Need to highlight specific clades or features
- Integrate additional data (traits, geography, expression)
:::
```{=html}
<!-- 
slide 62
-->
```

## Basic Tree Layouts

:::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 6
#| message: false
#| warning: false

library(ggtree)
library(ape)

# Create example tree
set.seed(123)
simple_tree <- rtree(15)

# Basic rectangular layout
ggtree(simple_tree) + 
  geom_tiplab(size = 3) +
  labs(title = "Rectangular Layout") +
  theme_tree2()
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 6

# Circular layout
ggtree(simple_tree, layout = "circular") + 
  geom_tiplab(size = 3) +
  labs(title = "Circular Layout")
```
:::
::::::

**Most common:** Rectangular (left) for detailed analysis, Circular (right) for compact display
```{=html}
<!-- 
slide 63
-->
```

## ggtree: Basic Tree Plotting

**Simple tree visualization:**
```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "1-2|4-5|7-10"

library(ggtree)
library(ape)

# Load or create a tree
tree <- read.tree("your_tree.nwk")

# Basic plot
ggtree(tree) + 
  geom_tiplab(size = 3) +
  theme_tree2()
```

::: fragment
**Common layouts:**

- `layout = "rectangular"` (default)
- `layout = "circular"`
- `layout = "fan"` (semicircular)
- `layout = "unrooted"`
:::
```{=html}
<!-- 
slide 64
-->
```

## Simple Tree Examples

:::::: columns
::: {.column width="50%"}
**Flipped horizontal:**
```{r}
#| echo: true
#| eval: false

ggtree(tree) + 
  coord_flip() +
  geom_tiplab(size = 3)
```
```{r}
#| echo: false
#| fig-width: 4.5
#| fig-height: 5

set.seed(123)
simple_tree <- rtree(12)
ggtree(simple_tree) + 
  coord_flip() +
  geom_tiplab(size = 3)
```
:::

::: {.column width="50%"}
**Reversed direction:**
```{r}
#| echo: true
#| eval: false

ggtree(tree) + 
  coord_flip() + 
  scale_x_reverse() +
  geom_tiplab(size = 3)
```
```{r}
#| echo: false
#| fig-width: 4.5
#| fig-height: 5

ggtree(simple_tree) + 
  coord_flip() + 
  scale_x_reverse() +
  geom_tiplab(size = 3, hjust = 1)
```
:::
::::::
```{=html}
<!-- 
slide 65
-->
```

## Circular and Unrooted Layouts

:::::: columns
::: {.column width="50%"}
**Circular layout:**
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 5

set.seed(456)
circ_tree <- rtree(20)

cross_colors <- met.brewer("Cross", 5)

ggtree(circ_tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 0.5) +
  labs(title = "Circular Tree")
```
:::

::: {.column width="50%"}
**Unrooted layout:**
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 5

ggtree(circ_tree, layout = "unrooted") +
  geom_tiplab(size = 2.5) +
  labs(title = "Unrooted Tree")
```
:::
::::::

**Use circular** for large trees with space constraints  
**Use unrooted** when root position is uncertain or irrelevant
```{=html}
<!-- 
slide 66
-->
```

## Highlighting Clades

**Emphasize specific groups:**
```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "1-3|5-7"

# Highlight a clade
ggtree(tree, layout = "circular") +
  geom_hilight(node = 23, fill = cross_colors[1], alpha = 0.6) +
  geom_tiplab(size = 3)

# Highlight multiple clades
ggtree(tree, layout = "circular") +
  geom_hilight(node = 23, fill = cross_colors[1], alpha = 0.5, extend = 0.2) +
  geom_hilight(node = 45, fill = cross_colors[3], alpha = 0.5, extend = 0.2) +
  geom_tiplab(size = 3)
```

**Parameters:**
- `node`: Internal node number to highlight
- `fill`: Color for the highlight
- `alpha`: Transparency (0-1)
- `extend`: How far to extend the highlight
```{=html}
<!-- 
slide 67
-->
```

## Example: Highlighted Clades
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| fig-align: center

set.seed(789)
large_tree <- rtree(50)

cross_colors <- met.brewer("Cross", 5)

# Find some internal nodes for highlighting
p <- ggtree(large_tree, layout = "circular")

# Highlight different clades
p + 
  geom_hilight(node = 55, fill = cross_colors[1], alpha = 0.4, extend = 0.15) +
  geom_hilight(node = 70, fill = cross_colors[3], alpha = 0.4, extend = 0.15) +
  geom_hilight(node = 85, fill = cross_colors[5], alpha = 0.4, extend = 0.15) +
  geom_tiplab(size = 2, align = FALSE) +
  labs(title = "Phylogenetic Tree with Highlighted Clades",
       subtitle = "Different colors represent distinct evolutionary groups")
```
```{=html}
<!-- 
slide 68
-->
```

## Adding Strips and Annotations

**Mark specific groups with colored strips:**
```{r}
#| echo: true
#| eval: false

ggtree(tree, layout = "circular") +
  geom_strip(
    taxa1 = "Species_A", 
    taxa2 = "Species_J",
    color = cross_colors[1], 
    barsize = 3,
    offset = 0.05
  ) +
  geom_tiplab(size = 3)
```

::: fragment
**Useful for:**

- Marking geographic groups
- Highlighting experimental groups
- Showing taxonomic divisions
- Indicating gene presence/absence
:::
```{=html}
<!-- 
slide 69
-->
```

## Fancy Tree: Multiple Annotations
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(ggtree)
library(ggtreeExtra)

set.seed(101)
fancy_tree <- rtree(40)

# Create some example data for the tips
tip_data <- data.frame(
  taxa = fancy_tree$tip.label,
  group = sample(c("Group A", "Group B", "Group C", "Group D"), 
                 40, replace = TRUE),
  value1 = rnorm(40, mean = 50, sd = 15),
  value2 = runif(40, 0, 100)
)

cross_colors <- met.brewer("Cross", 5)

p <- ggtree(fancy_tree, layout = "circular", size = 0.8)

# Add highlighted clades
p2 <- p + 
  geom_hilight(node = 45, fill = cross_colors[1], alpha = 0.3, extend = 0.1) +
  geom_hilight(node = 60, fill = cross_colors[2], alpha = 0.3, extend = 0.1) +
  geom_hilight(node = 75, fill = cross_colors[4], alpha = 0.3, extend = 0.1)

# Add tip labels
p3 <- p2 + 
  geom_tiplab(size = 2, align = TRUE, linesize = 0.3, offset = 0.5) +
  geom_tippoint(aes(color = tip_data$group[match(label, tip_data$taxa)]), 
                size = 2, alpha = 0.8) +
  scale_color_manual(values = cross_colors[1:4], name = "Group")

# Add title
p3 + 
  labs(title = "Annotated Phylogenetic Tree",
       subtitle = "With highlighted clades and tip metadata") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 12, hjust = 0.5))
```
```{=html}
<!-- 
slide 70
-->
```

## Adding External Data with ggtreeExtra

**Heatmaps alongside trees:**
```{r}
#| echo: true
#| eval: false

library(ggtreeExtra)

# Tree with heatmap
ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2, offset = 0.5) +
  geom_fruit(
    data = trait_data,
    geom = geom_tile,
    mapping = aes(y = taxa, x = trait, fill = value),
    offset = 0.05,
    pwidth = 0.1
  ) +
  scale_fill_gradientn(colors = met.brewer("Cross", 5))
```

**Add multiple data layers:**
- Heatmaps (gene expression, traits)
- Bar plots (abundance, counts)
- Points (presence/absence)
- Strips (categorical data)
```{=html}
<!-- 
slide 71
-->
```

## Example: Tree with Heatmap
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 7
#| fig-align: center
#| message: false
#| warning: false

library(ggtreeExtra)

set.seed(202)
heat_tree <- rtree(25)

# Create heatmap data
heat_data <- expand.grid(
  taxa = heat_tree$tip.label,
  trait = paste0("Trait", 1:5)
)
heat_data$value <- rnorm(nrow(heat_data), mean = 50, sd = 20)

cross_colors <- met.brewer("Cross", 5)

ggtree(heat_tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 1.5) +
  geom_fruit(
    data = heat_data,
    geom = geom_tile,
    mapping = aes(y = taxa, x = trait, fill = value),
    offset = 0.1,
    pwidth = 0.25,
    color = "white",
    size = 0.5
  ) +
  scale_fill_gradientn(
    colors = c(cross_colors[1], "white", cross_colors[3]),
    name = "Value"
  ) +
  labs(title = "Phylogenetic Tree with Trait Heatmap",
       subtitle = "Integrating evolutionary relationships with trait data") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 12, hjust = 0.5))
```
```{=html}
<!-- 
slide 72
-->
```

## Tree Comparison: Tanglegram

**Compare two trees side-by-side:**
```{r}
#| echo: true
#| eval: false

library(ggtree)

# Create two tree plots
p1 <- ggtree(tree1) + geom_tiplab()
p2 <- ggtree(tree2) + geom_tiplab()

# Combine for comparison
p1 | p2
```

::: fragment
**Use for:**

- Compare gene trees vs species trees
- Before/after tree optimization
- Different phylogenetic methods
- Geographic vs genetic relationships
:::
```{=html}
<!-- 
slide 73
-->
```

## Best Practices for Tree Visualization

**Design tips:**

::: fragment
**1. Choose layout wisely**

- Small trees (<50 tips): Rectangular with labels
- Large trees (50-200): Circular or fan
- Very large (>200): Consider collapsing clades
:::

::: fragment
**2. Use color strategically**

- Highlight functional groups or clades
- Color by metadata (geography, phenotype)
- Use consistent color schemes across figures
:::

::: fragment
**3. Integrate data thoughtfully**

- Add heatmaps for quantitative traits
- Use symbols for categorical data
- Don't overcrowd with too many layers
:::
```{=html}
<!-- 
slide 74
-->
```

## Interactive Trees with ggtree

**Export for interactive exploration:**
```{r}
#| echo: true
#| eval: false

library(plotly)

# Create tree
p <- ggtree(tree) + geom_tiplab(size = 3)

# Convert to interactive
ggplotly(p)
```

::: fragment
**Advantages:**

- Zoom into specific clades
- Hover for tip information
- Toggle layers on/off
- Suitable for web reports and supplementary materials
:::
```{=html}
<!-- 
slide 75
-->
```

## Phylogenetic Tree Packages Comparison
```{r}
#| echo: false

library(gt)

tree_packages <- tibble(
  Package = c("ggtree", "ape", "phytools", "treeio", "ggtreeExtra"),
  Purpose = c(
    "Tree visualization with ggplot2",
    "Phylogenetic analysis & basic plots",
    "Comparative methods & plotting",
    "Tree data import/export",
    "Add external data layers to ggtree"
  ),
  Strengths = c(
    "Flexible, ggplot2 integration",
    "Foundation for tree analysis",
    "Ancestral state reconstruction",
    "Multiple tree format support",
    "Easy data integration"
  ),
  `Best For` = c(
    "Publication-quality figures",
    "Tree manipulation & analysis",
    "Evolutionary modeling",
    "Data import/conversion",
    "Multi-layered annotations"
  )
)

tree_packages |>
  gt() |>
  cols_align(align = "left") |>
  tab_style(
    style = cell_text(weight = "bold", size = px(14)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(size = px(12)),
    locations = cells_body()
  ) |>
  tab_options(
    table.font.size = px(13),
    table.width = pct(95)
  )
```
