---
title: "Data Visualization - day2"
subtitle: "<span style='font-size:1.25em;'>for Genomics and Biological Sciences</span>"
author: "<div style='width: 100%; text-align: center; margin-left: auto; margin-right: auto;'><a href='https://github.com/loukesio' style='color:#28a87d;font-weight:600;'>Dr. Loukas Theodosiou</a>&ensp;//&ensp;December 15<sup>th</sup> 2025</div>"
title-slide-attributes:
    data-background-color: "#E8F5F1"
format:
  revealjs:
    theme: slides.scss
    footer: '**L.Theodosiou** // Consultation in Genomics Data Visualization'
    transition: fade
    progress: true
    multiplex: false
    preview-links: true
    scrollable: false
    hide-inactive-cursor: true
    highlight-style: printing
    pause: true
    embed-resources: false
editor: source
knitr:
  opts_chunk:
    dev: "ragg_png"
    retina: 3
    dpi: 200
execute:
  freeze: auto
  cache: true
  echo: true
  fig-width: 12
  fig-height: 7

---

```{r}
#| label: setup
#| include: false

# Core tidyverse packages
library(tidyverse)      # Includes dplyr, ggplot2, tidyr, readr, purrr, stringr

# Plot composition
library(patchwork)      # Combine multiple plots
library(cowplot)        # Additional plot arrangement tools

# Color palettes and tables
library(MetBrewer)      # Color palettes from art
library(RColorBrewer)   # Additional color palettes
library(viridis)        # Colorblind-friendly palettes
library(scales)         # Scale functions for ggplot2
library(colorspace)     # Color manipulation
library(gt)             # Publication-quality tables

# Specialized ggplot2 extensions
library(ggrepel)        # Non-overlapping text labels
library(ggforce)        # Additional ggplot2 features
library(ggridges)       # Ridge plots

# Pathway enrichment visualization
library(clusterProfiler)  # GO/KEGG enrichment (Bioconductor)
library(enrichplot)       # Enrichment visualization (Bioconductor)
library(ReactomePA)       # Reactome pathway analysis (Bioconductor)

# Heatmaps
library(pheatmap)       # Pretty heatmaps
library(ComplexHeatmap) # Advanced heatmaps (Bioconductor)

# Set comparisons
library(VennDiagram)    # Venn diagrams
library(UpSetR)         # UpSet plots for complex overlaps

# Dimensionality reduction
library(umap)           # UMAP
library(Rtsne)          # t-SNE

# Genomics visualization - Circular plots
library(circlize)       # Circular layout plots
library(OmicCircos)     # Multi-omics circular plots (Bioconductor)

# Genomics visualization - Gene arrangements
library(gggenes)        # Gene arrow maps
library(genoPlotR)      # Genome comparisons

# Genomics visualization - Chromosomes
library(GenomicRanges)  # Genomic intervals (Bioconductor)
library(IRanges)        # Integer ranges (Bioconductor)
library(karyoploteR)    # Karyotype plots (Bioconductor)
library(Gviz)           # Genomic tracks (Bioconductor)
library(chromoMap)      # Interactive chromosome maps
library(RIdeogram)      # Chromosome ideograms

# Phylogenetic trees
library(ape)            # Phylogenetic analysis
library(ggtree)         # Tree visualization with ggplot2 (Bioconductor)
library(treeio)         # Tree data import/export (Bioconductor)
library(phytools)       # Phylogenetic tools
library(ggtreeExtra)    # ggtree extensions (Bioconductor)

# Network and graph visualization
library(tidygraph)      # Tidy graph manipulation
library(igraph)         # Network analysis
library(ggraph)         # Network visualization with ggplot2

# Interactive plots
library(plotly)         # Interactive plots

# Custom fonts (optional)
library(showtext)       # Google Fonts integration
# Uncomment to use custom fonts:
# font_add_google("Roboto", "roboto")
# showtext_auto()

# Define a custom theme function for ggplot2 to enhance plot aesthetics
custom_theme <- function(base_size = 12, 
                         title_size_rel = 1.5,
                         subtitle_size_rel = 1.25,
                         axis_title_size_rel = 1.25, 
                         axis_text_size_rel = 1.25, 
                         strip_text_size_rel = 1.1) {
  # Use the theme_bw base theme with a specified font family ("Asap")
  theme_bw(base_size = base_size, base_family = "Asap") +
    theme(
      # Customize the title: center-align and set relative size
      plot.title = element_text(hjust = 0.5, size = rel(title_size_rel)),
      # Customize the subtitle: center-align and set relative size
      plot.subtitle = element_text(hjust = 0.5, size = rel(subtitle_size_rel)),
      # Customize the x-axis title: set size and add a top margin
      axis.title.x = element_text(size = rel(axis_title_size_rel), 
                                  margin = margin(t = 12.5)),
      # Customize the y-axis title: set size and add a right margin
      axis.title.y = element_text(size = rel(axis_title_size_rel), 
                                  margin = margin(r = 12.5)),
      # Customize the x-axis text: set size
      axis.text.x = element_text(size = rel(axis_text_size_rel)),
      # Customize the y-axis text: set size
      axis.text.y = element_text(size = rel(axis_text_size_rel)),
      # Customize facet strip text size
      strip.text = element_text(size = rel(strip_text_size_rel)),
      # Add margins around the plot t,r,b,l
      plot.margin = unit(c(2, 2, 1, 1), "cm"),
      # Remove major grid lines on the x-axis
      panel.grid.major.x = element_blank(),
      panel.grid.major.y  = element_blank(),
      # Remove minor grid lines on x-axis
      panel.grid.minor.x = element_blank(),
      # Keep minor grid lines on y-axis as dashed
      panel.grid.minor.y = element_line(
        linetype = "dashed",
        linewidth = 0.3,
        colour = "grey80"
      ),
      # Remove background from facet strips
      strip.background = element_blank()
    )
}

# Set custom theme as default
theme_set(custom_theme())

# Set default color palette
cross_colors <- met.brewer("Cross", 5)
```

```{=html}

<!--
   _____                                 _                     
  / ____|                               (_)                    
 | |     ___  _ __ ___  _ __   __ _ _ __ _ ___  ___  _ __  ___ 
 | |    / _ \| '_ ` _ \| '_ \ / _` | '__| / __|/ _ \| '_ \/ __|
 | |___| (_) | | | | | | |_) | (_| | |  | \__ \ (_) | | | \__ \
  \_____\___/|_| |_| |_| .__/ \__,_|_|  |_|___/\___/|_| |_|___/
                       | |                                     
                       |_|            
-->

```

# Set comparisons: Venn & UpSet {background-color="#28A87D"}
```{=html}
<!-- 
slide venn2
-->
```

## Why Compare Gene Sets?

:::::: columns
::: {.column width="50%"}
**Common scenarios:**

- **Differential expression**: Which genes change across conditions?
- **Pathway analysis**: Which genes overlap between pathways?
- **Multi-omics**: Which features appear in multiple datasets?
- **Comparative genomics**: Which genes are shared/unique?

::: fragment
**The challenge:**

How do you visualize overlaps between 3, 4, 5+ sets clearly?
:::
:::

::: {.column width="50%"}
::: fragment
**Two approaches:**

**Venn diagrams:**
- ✅ Intuitive for 2-3 sets
- ❌ Messy for 4+ sets
- Shows all possible intersections

**UpSet plots:**
- ✅ Scales to many sets
- ✅ Shows intersection sizes clearly
- ✅ More informative than Venn
:::
:::
::::::
```{=html}
<!-- 
slide venn3
-->
```

## When to Use Each Method
```{r}
#| echo: false

library(gt)

comparison <- tibble(
  Feature = c("Best for", "Maximum sets", "Shows sizes", 
              "Easy to read", "Quantitative", "Interpretation"),
  `Venn Diagram` = c(
    "2-3 sets",
    "~4 sets (becomes unclear)",
    "Yes (with labels)",
    "High (for 2-3 sets)",
    "Circles not proportional",
    "Quick overview"
  ),
  `UpSet Plot` = c(
    "3+ sets",
    "10+ sets easily",
    "Yes (bar heights)",
    "High (any number)",
    "Exact counts shown",
    "Detailed comparisons"
  )
)

comparison |>
  gt() |>
  cols_align(align = "left") |>
  tab_style(
    style = cell_text(weight = "bold", size = px(14)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(size = px(12)),
    locations = cells_body()
  ) |>
  tab_options(
    table.font.size = px(13),
    table.width = pct(95)
  )
```

**Rule of thumb:** Use Venn for ≤3 sets, UpSet for ≥4 sets
```{=html}
<!-- 
slide venn4
-->
```

## Example Dataset: Gene Lists

**Create gene set data:**
```{r}
#| echo: true
#| eval: false

# Human genes from different experiments
human_genes <- c(
  "BRCA1", "TP53", "EGFR", "VEGFA", "MYC", "PTEN", "CDKN2A", "RB1", "APC", "KRAS",
  "BRAF", "ATM", "ABL1", "JAK2", "SMAD4", "FGFR3", "PIK3CA", "NTRK1", "CTNNB1", "HNF1A",
  "GATA3", "RUNX1", "MLH1", "PMS2", "MSH2", "MSH6", "PDGFRA", "ERBB2", "MPL", "NOTCH1",
  "IDH1", "IDH2", "KIT", "FLT3", "CSF1R", "SRC", "CDH1", "MET", "ALK", "ROS1",
  "RET", "SMO", "FOXA1", "AR", "WT1", "NF1", "TSC1", "TSC2", "STK11", "VHL",
  "SDHB", "SDHD", "NFKB1", "STAT3", "ELANE", "CEBPA", "KLF4", "FOXP3", "SOX2", "TBX3",
  "TBX5", "HNF4A", "PPARG", "IRF4", "IRF8", "ZEB1", "ZEB2", "EZH2", "SMARCB1", "YAP1",
  "TAZ", "TEAD1", "TEAD4", "FOXM1", "E2F1", "E2F3", "MMP9", "MMP2", "BCL2", "BAX",
  "BID", "BAK1", "CYC1", "COX6A2", "COX7A2L", "ATP5F1", "ATP5MC1", "ATP5MC3", "ATP5PF", "ATP5PO",
  "ATP5SL", "ATP5MG", "NDUFA4", "NDUFB2", "NDUFB4", "NDUFB5", "NDUFB6", "NDUFB7", "NDUFB8", "NDUFS1",
  "NDUFS2", "NDUFS3", "NDUFS5", "NDUFS6", "NDUFV1", "NDUFV2", "NDUFV3", "SDHA", "SDHC", "UQCRC1"
)

# Sample 100 genes per dataset (simulating different experiments)
set.seed(123)
gene_sets_3 <- list(
  Experiment_1 = sample(human_genes, 100, replace = FALSE),
  Experiment_2 = sample(human_genes, 100, replace = FALSE),
  Experiment_3 = sample(human_genes, 100, replace = FALSE)
)

gene_sets_5 <- list(
  Experiment_1 = sample(human_genes, 100, replace = FALSE),
  Experiment_2 = sample(human_genes, 100, replace = FALSE),
  Experiment_3 = sample(human_genes, 100, replace = FALSE),
  Experiment_4 = sample(human_genes, 100, replace = FALSE),
  Experiment_5 = sample(human_genes, 100, replace = FALSE)
)
```

**Data structure:** Named list where each element is a vector of gene names
```{=html}
<!-- 
slide venn5
-->
```

## Venn Diagram: Method 1 (VennDiagram)

**Traditional approach:**
```{r}
#| echo: true
#| eval: false

library(VennDiagram)
library(grid)

# Create Venn diagram
venn_plot <- venn.diagram(
  x = gene_sets_3,
  filename = NULL,  # Don't save to file
  col = c("#440154ff", "#21908dff", "#fde725ff"),
  fill = c(alpha("#440154ff", 0.3), 
           alpha("#21908dff", 0.3), 
           alpha("#fde725ff", 0.3)),
  cex = 1.5,              # Label size
  cat.cex = 1.2,          # Category name size
  cat.fontface = "bold"
)

# Display it
grid.newpage()
grid.draw(venn_plot)
```

**Note:** Returns a grid object, need `grid.draw()` to display
```{=html}
<!-- 
slide venn5a
-->
```

## Venn Diagram Output (VennDiagram)
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
#| fig-align: center
#| warning: false
#| message: false

library(VennDiagram)
library(grid)

# Use the gene sets from setup
human_genes <- c(
  "BRCA1", "TP53", "EGFR", "VEGFA", "MYC", "PTEN", "CDKN2A", "RB1", "APC", "KRAS",
  "BRAF", "ATM", "ABL1", "JAK2", "SMAD4", "FGFR3", "PIK3CA", "NTRK1", "CTNNB1", "HNF1A",
  "GATA3", "RUNX1", "MLH1", "PMS2", "MSH2", "MSH6", "PDGFRA", "ERBB2", "MPL", "NOTCH1",
  "IDH1", "IDH2", "KIT", "FLT3", "CSF1R", "SRC", "CDH1", "MET", "ALK", "ROS1",
  "RET", "SMO", "FOXA1", "AR", "WT1", "NF1", "TSC1", "TSC2", "STK11", "VHL",
  "SDHB", "SDHD", "NFKB1", "STAT3", "ELANE", "CEBPA", "KLF4", "FOXP3", "SOX2", "TBX3",
  "TBX5", "HNF4A", "PPARG", "IRF4", "IRF8", "ZEB1", "ZEB2", "EZH2", "SMARCB1", "YAP1",
  "TAZ", "TEAD1", "TEAD4", "FOXM1", "E2F1", "E2F3", "MMP9", "MMP2", "BCL2", "BAX",
  "BID", "BAK1", "CYC1", "COX6A2", "COX7A2L", "ATP5F1", "ATP5MC1", "ATP5MC3", "ATP5PF", "ATP5PO",
  "ATP5SL", "ATP5MG", "NDUFA4", "NDUFB2", "NDUFB4", "NDUFB5", "NDUFB6", "NDUFB7", "NDUFB8", "NDUFS1",
  "NDUFS2", "NDUFS3", "NDUFS5", "NDUFS6", "NDUFV1", "NDUFV2", "NDUFV3", "SDHA", "SDHC", "UQCRC1"
)

set.seed(123)
gene_sets_3 <- list(
  Experiment_1 = sample(human_genes, 100, replace = FALSE),
  Experiment_2 = sample(human_genes, 100, replace = FALSE),
  Experiment_3 = sample(human_genes, 100, replace = FALSE)
)

venn_plot <- venn.diagram(
  x = gene_sets_3,
  filename = NULL,
  col = c("#440154ff", "#21908dff", "#fde725ff"),
  fill = c(alpha("#440154ff", 0.3), 
           alpha("#21908dff", 0.3), 
           alpha("#fde725ff", 0.3)),
  cex = 2,
  cat.cex = 1.5,
  cat.fontface = "bold",
  cat.dist = c(0.05, 0.05, 0.05),
  margin = 0.1
)

grid.newpage()
grid.draw(venn_plot)
```

**Interpretation:** Numbers show gene counts in each region
```{=html}
<!-- 
slide venn6
-->
```

## Venn Diagram: Method 2 (ggvenn)

**ggplot2-compatible approach:**
```{r}
#| echo: true
#| eval: false

library(ggvenn)

# Much simpler syntax!
ggvenn(
  gene_sets_3,
  fill_color = c("#0073C2FF", "#EFC000FF", "#868686FF", "#CD534CFF"),
  stroke_size = 0.5,
  set_name_size = 5,
  text_size = 4
)
```

**Advantages:**
- ✅ Cleaner syntax
- ✅ Works with ggplot2
- ✅ Easy to customize
- ✅ No need for `grid.draw()`
```{=html}
<!-- 
slide venn6a
-->
```

## Venn Diagram Output (ggvenn)
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
#| fig-align: center
#| warning: false
#| message: false

library(ggvenn)

ggvenn(
  gene_sets_3,
  fill_color = c("#0073C2FF", "#EFC000FF", "#868686FF"),
  stroke_size = 0.5,
  set_name_size = 6,
  text_size = 5
) +
  labs(title = "Gene Overlap Across Three Experiments") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 16))
```
```{=html}
<!-- 
slide venn7
-->
```

## Venn with ggVennDiagram

**Most modern approach with ggplot2:**
```{r}
#| echo: true
#| eval: false

library(ggVennDiagram)

# Create Venn with set labels and percentages
ggVennDiagram(
  gene_sets_3,
  label = "count",           # Show counts
  label_alpha = 0,           # Transparent label background
  edge_size = 1.5            # Border thickness
) +
  scale_fill_gradient(low = "white", high = cross_colors[1]) +
  labs(title = "Gene Set Overlaps",
       fill = "Count") +
  theme(legend.position = "right")
```

**Features:**
- Shows counts or percentages
- Color by overlap size
- Full ggplot2 customization
```{=html}
<!-- 
slide venn7a
-->
```

## ggVennDiagram Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| fig-align: center
#| warning: false
#| message: false

library(ggVennDiagram)

cross_colors <- met.brewer("Cross", 5)

ggVennDiagram(
  gene_sets_3,
  label = "count",
  label_alpha = 0,
  label_size = 5,
  edge_size = 1.5
) +
  scale_fill_gradient(low = "white", high = cross_colors[1]) +
  labs(title = "Gene Set Overlaps Across Experiments",
       subtitle = "Color intensity shows overlap size",
       fill = "Gene\nCount") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "right"
  )
```
```{=html}
<!-- 
slide venn8
-->
```

## Problem: Too Many Sets

**What happens with 5 sets?**
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| fig-align: center
#| warning: false
#| message: false

set.seed(123)
gene_sets_5 <- list(
  Exp_1 = sample(human_genes, 100, replace = FALSE),
  Exp_2 = sample(human_genes, 100, replace = FALSE),
  Exp_3 = sample(human_genes, 100, replace = FALSE),
  Exp_4 = sample(human_genes, 100, replace = FALSE),
  Exp_5 = sample(human_genes, 100, replace = FALSE)
)

venn_plot_5 <- venn.diagram(
  x = gene_sets_5,
  filename = NULL,
  col = c("#440154ff", "#21908dff", "#fde725ff", "#b73779", "#3b528b"),
  fill = c(alpha("#440154ff", 0.3), 
           alpha("#21908dff", 0.3), 
           alpha("#fde725ff", 0.3),
           alpha("#b73779", 0.3),
           alpha("#3b528b", 0.3)),
  cex = 1.2,
  cat.cex = 1,
  margin = 0.05
)

grid.newpage()
grid.draw(venn_plot_5)
```

**Problem:** Cluttered, hard to interpret, some regions unlabeled

**Solution:** UpSet plots! →
```{=html}
<!-- 
slide venn9
-->
```

## UpSet Plots: Better for Multiple Sets

**Code:**
```{r}
#| echo: true
#| eval: false

library(UpSetR)

# Convert list to binary matrix
upset_data <- fromList(gene_sets_5)

# Create UpSet plot
upset(
  upset_data,
  nsets = 5,                    # Number of sets to show
  order.by = "freq",            # Order by frequency
  main.bar.color = cross_colors[1],
  sets.bar.color = cross_colors[3],
  text.scale = 1.5,             # Increase text size
  point.size = 4,
  line.size = 1.5
)
```

**Key concepts:**
- **Bottom bars**: Size of each set
- **Top bars**: Size of each intersection
- **Dots + lines**: Which sets contribute to intersection
```{=html}
<!-- 
slide venn9a
-->
```

## UpSet Plot Output
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 7
#| fig-align: center
#| warning: false
#| message: false

library(UpSetR)

upset_data_5 <- fromList(gene_sets_5)

cross_colors <- met.brewer("Cross", 5)

upset(
  upset_data_5,
  nsets = 5,
  order.by = "freq",
  main.bar.color = cross_colors[1],
  sets.bar.color = cross_colors[3],
  text.scale = c(1.5, 1.5, 1.3, 1.3, 1.5, 1.5),
  point.size = 4,
  line.size = 1.5,
  mainbar.y.label = "Intersection Size",
  sets.x.label = "Set Size"
)
```

**Reading:** Tallest bar = largest intersection (genes in Exp_1 AND Exp_2 only)
```{=html}
<!-- 
slide venn10
-->
```

## How to Read UpSet Plots

:::::: columns
::: {.column width="50%"}
**Bottom section:**

- Horizontal bars = set sizes
- Shows total genes per experiment

**Top section:**

- Vertical bars = intersection sizes
- Height = number of genes
:::

::: {.column width="50%"}
**Connection matrix:**

- **Single dot**: Unique to one set
- **Multiple dots + line**: Shared by those sets
- **Empty**: Not in that set

**Example:**
```
● ━ ●     = Exp_1 AND Exp_3 only
●         = Exp_1 only  
● ━ ● ━ ● = All three
```
:::
::::::
```{=html}
<!-- 
slide venn11
-->
```

## Exercise 1: Differential Expression Venn

::: {.question}
**Compare DEGs across tissues:**
```{r}
#| echo: true
#| eval: false

# Simulated differential expression results
set.seed(456)

all_genes <- paste0("Gene_", sprintf("%04d", 1:2000))

deg_sets <- list(
  Brain = sample(all_genes, 150),
  Liver = sample(all_genes, 200),
  Heart = sample(all_genes, 120),
  Kidney = sample(all_genes, 180)
)

# Your tasks:
# 1. Create Venn diagram with ggvenn (3 tissues only: Brain, Liver, Heart)
# 2. Use Cross color palette
# 3. Add appropriate title
# 4. Create UpSet plot with all 4 tissues
# 5. Order by frequency (order.by = "freq")
# 6. Identify which intersection is largest

# Questions to answer:
# - How many genes are unique to Brain?
# - How many genes are shared by all three tissues?
# - Which tissue has the most DEGs?
```
:::
```{=html}
<!-- 
slide venn12
-->
```

## Exercise 2: Pathway Overlap Analysis

::: {.question}
**Analyze genes from 5 biological pathways:**
```{r}
#| echo: true
#| eval: false

# Simulated pathway gene lists
set.seed(789)

all_genes <- c(paste0("Gene_", 1:500))

pathway_genes <- list(
  "PI3K-Akt signaling" = sample(all_genes, 80),
  "MAPK signaling" = sample(all_genes, 75),
  "Cell cycle" = sample(all_genes, 60),
  "p53 pathway" = sample(all_genes, 45),
  "Apoptosis" = sample(all_genes, 55)
)

# Your tasks:
# 1. Create UpSet plot showing all pathway overlaps
# 2. Color main bars with cross_colors[1]
# 3. Color set bars with cross_colors[3]
# 4. Order intersections by frequency
# 5. Increase text size for readability (text.scale = 1.5)

# Advanced:
# 6. Try ggVennDiagram with 3 pathways (PI3K-Akt, MAPK, Cell cycle)
# 7. Use gradient fill based on overlap size
# 8. Extract genes in the largest intersection

# Hint for extraction:
# Use Reduce(intersect, list(set1, set2, set3))
```
:::
```{=html}
<!-- 
slide venn13
-->
```

## Extracting Specific Intersections

**Get genes from specific overlaps:**
```{r}
#| echo: true
#| eval: false

# Intersection of all sets
common_all <- Reduce(intersect, gene_sets_3)
length(common_all)  # How many?
head(common_all)    # Which genes?

# Only in Experiment_1 and Experiment_2 (not in 3)
exp1_exp2_only <- setdiff(
  intersect(gene_sets_3$Experiment_1, gene_sets_3$Experiment_2),
  gene_sets_3$Experiment_3
)

# Unique to Experiment_1
unique_exp1 <- setdiff(
  gene_sets_3$Experiment_1,
  c(gene_sets_3$Experiment_2, gene_sets_3$Experiment_3)
)

# All genes in any set (union)
all_genes <- Reduce(union, gene_sets_3)
```
```{=html}
<!-- 
slide venn14
-->
```

## Comparison: All Three Methods
```{r}
#| echo: false

library(gt)

method_comparison <- tibble(
  Method = c("VennDiagram", "ggvenn", "ggVennDiagram", "UpSetR"),
  `Syntax` = c("Complex", "Simple", "Simple", "Medium"),
  `ggplot2` = c("❌", "✅", "✅", "❌"),
  `Max Sets` = c("~4", "~4", "~4", "10+"),
  `Customization` = c("High", "Medium", "High", "High"),
  `Best For` = c(
    "Traditional Venn",
    "Quick 2-3 set Venn",
    "Modern styled Venn",
    "Many sets (4+)"
  )
)

method_comparison |>
  gt() |>
  cols_align(align = "left") |>
  tab_style(
    style = cell_text(weight = "bold", size = px(14)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(size = px(12)),
    locations = cells_body()
  ) |>
  tab_options(
    table.font.size = px(13),
    table.width = pct(95)
  )
```

**Recommendation:** Use ggVennDiagram for ≤3 sets, UpSetR for ≥4 sets
```{=html}
<!-- 
slide venn15
-->
```

## Best Practices

**Design guidelines:**

::: fragment
**For Venn diagrams:**

✅ Use for 2-3 sets maximum  
✅ Show counts/numbers in regions  
✅ Use transparent fills (alpha = 0.3)  
✅ Choose contrasting colors  
❌ Don't use for >4 sets
:::

::: fragment
**For UpSet plots:**

✅ Order by frequency (most common first)  
✅ Increase text size for readability  
✅ Use consistent colors  
✅ Show set sizes on bottom  
✅ Limit to top 20-30 intersections
:::

::: fragment
**General:**

- Always label your sets clearly
- Include sample sizes in title/caption
- Export at high resolution for publication
:::


```{=html}
<!-- 
slide tree1
-->
```

# Phylogenetic trees {background-color="#28A87D"}
```{=html}
<!-- 
slide tree2
-->
```

## What are Phylogenetic Trees?

:::::: columns
::: {.column width="50%"}
**Definition:**

Trees represent evolutionary relationships between species, genes, or sequences

::: fragment
**Key components:**

- **Tips/leaves**: Species or genes
- **Branches**: Evolutionary paths
- **Nodes**: Common ancestors
- **Branch lengths**: Evolutionary distance
:::
:::

::: {.column width="50%"}
::: fragment
**Why visualize trees?**

- Show evolutionary relationships
- Identify gene families
- Compare species divergence
- Analyze trait evolution
- Detect horizontal gene transfer

**Common in:**

- Comparative genomics
- Population genetics
- Molecular evolution
- Phylogeography
:::
:::
::::::
```{=html}
<!-- 
slide tree3
-->
```

## Tree Layouts: Choose Your View
```{r}
#| echo: false

library(gt)

layout_comparison <- tibble(
  Layout = c("Rectangular", "Circular", "Radial", "Fan", "Unrooted"),
  `Best For` = c(
    "Small-medium trees (< 50 tips)",
    "Large trees, aesthetic",
    "Medium trees, space-efficient",
    "Large trees, semicircular",
    "Networks, no root"
  ),
  `Shows Time` = c("Yes", "No", "No", "Partial", "No"),
  `Space Usage` = c("Linear", "Compact", "Compact", "Medium", "Variable"),
  `Readability` = c("High", "Medium", "High", "Medium", "Low")
)

layout_comparison |>
  gt() |>
  cols_align(align = "left") |>
  tab_style(
    style = cell_text(weight = "bold", size = px(14)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(size = px(11)),
    locations = cells_body()
  ) |>
  tab_options(
    table.font.size = px(12),
    table.width = pct(95)
  )
```

**We'll explore all of these!**
```{=html}
<!-- 
slide tree4
-->
```

## Load Example Tree

**Step 1: Read tree file**
```{r}
#| echo: true
#| eval: false

library(ggtree)
library(ape)

# Read Newick format tree
tree <- read.tree("itol.nwk")

# Or generate random tree for practice
set.seed(123)
tree <- rtree(30)  # 30 tips

# Examine tree structure
tree
#   Number of tips: 30
#   Number of nodes: 29
#   Branch lengths: Yes
```

**Tree files contain:**
- Topology (relationships)
- Branch lengths (distances)
- Tip labels (species/gene names)
```{=html}
<!-- 
slide tree5
-->
```

## Basic Tree: Rectangular Layout

**Code:**
```{r}
#| echo: true
#| eval: false

library(ggtree)

# Most basic tree
ggtree(tree) +
  geom_tiplab(size = 3) +
  labs(title = "Basic Phylogenetic Tree")
```

**Default:**
- Rectangular layout
- Tips on right
- Root on left
- Black branches
```{=html}
<!-- 
slide tree5a
-->
```

## Basic Tree: Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center

library(ggtree)
library(ape)

set.seed(123)
tree <- rtree(30)

ggtree(tree) +
  geom_tiplab(size = 3) +
  labs(title = "Basic Rectangular Tree") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 16))
```
```{=html}
<!-- 
slide tree6
-->
```

## Flip and Reverse

**Horizontal flip:**
```{r}
#| echo: true
#| eval: false

# Flip horizontally
ggtree(tree) +
  coord_flip() +
  scale_x_reverse() +
  geom_tiplab(size = 3, hjust = 1) +
  labs(title = "Flipped Tree")
```

**When to use:**
- Match publication style
- Align with other figures
- Personal preference
```{=html}
<!-- 
slide tree6a
-->
```

## Flipped Tree Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center

ggtree(tree) +
  coord_flip() +
  scale_x_reverse() +
  geom_tiplab(size = 3, hjust = 1) +
  labs(title = "Horizontally Flipped Tree") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 16))
```
```{=html}
<!-- 
slide tree7
-->
```

## Circular Layout

**Code:**
```{r}
#| echo: true
#| eval: false

# Circular tree
ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 0.5) +
  labs(title = "Circular Phylogenetic Tree")
```

**Advantages:**
- Compact for large trees
- Aesthetically pleasing
- Good for posters/presentations

**Note:** Labels need `offset` to avoid overlap with branches
```{=html}
<!-- 
slide tree7a
-->
```

## Circular Tree Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 10
#| fig-align: center

ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 0.5) +
  labs(title = "Circular Layout") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 16))
```
```{=html}
<!-- 
slide tree8
-->
```

## Fan and Radial Layouts

:::::: columns
::: {.column width="50%"}
**Fan (semicircular):**
```{r}
#| echo: true
#| eval: false

ggtree(tree, layout = "fan") +
  geom_tiplab(size = 2.5)
```

- Half-circle
- Saves vertical space
- Good for wide figures
:::

::: {.column width="50%"}
**Radial:**
```{r}
#| echo: true
#| eval: false

ggtree(tree, layout = "radial") +
  geom_tiplab(size = 2.5)
```

- Radiates from center
- Similar to circular
- Different spacing
:::
::::::
```{=html}
<!-- 
slide tree8a
-->
```

## Fan and Radial Outputs
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
#| fig-align: center

library(patchwork)

p_fan <- ggtree(tree, layout = "fan") +
  geom_tiplab(size = 2.5) +
  labs(title = "Fan Layout") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p_radial <- ggtree(tree, layout = "radial") +
  geom_tiplab(size = 2.5, offset = 0.3) +
  labs(title = "Radial Layout") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p_fan | p_radial
```
```{=html}
<!-- 
slide tree9
-->
```

## Highlighting Clades

**Color specific groups:**
```{r}
#| echo: true
#| eval: false

# Identify nodes for your clades
# Node numbers above number of tips
clade1 <- 35
clade2 <- 45

ggtree(tree) +
  geom_hilight(node = clade1, 
               fill = cross_colors[1], 
               alpha = 0.3,
               extend = 0.2) +
  geom_hilight(node = clade2, 
               fill = cross_colors[3], 
               alpha = 0.3,
               extend = 0.2) +
  geom_tiplab(size = 3) +
  labs(title = "Highlighted Clades")
```

**Parameters:**
- `node`: Ancestor node number
- `fill`: Background color
- `alpha`: Transparency
- `extend`: How far to extend highlight
```{=html}
<!-- 
slide tree9a
-->
```

## Highlighted Clades Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center

cross_colors <- met.brewer("Cross", 5)

# Find suitable nodes
clade1 <- 35
clade2 <- 45

ggtree(tree) +
  geom_hilight(node = clade1, fill = cross_colors[1], alpha = 0.3, extend = 0.2) +
  geom_hilight(node = clade2, fill = cross_colors[3], alpha = 0.3, extend = 0.2) +
  geom_tiplab(size = 3) +
  labs(title = "Phylogenetic Tree with Highlighted Clades",
       subtitle = "Colors distinguish major evolutionary groups") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
        plot.subtitle = element_text(hjust = 0.5, size = 12))
```
```{=html}
<!-- 
slide tree10
-->
```

## Clade Labels

**Add text labels to clades:**
```{r}
#| echo: true
#| eval: false

ggtree(tree) +
  geom_hilight(node = clade1, fill = cross_colors[1], alpha = 0.3) +
  geom_hilight(node = clade2, fill = cross_colors[3], alpha = 0.3) +
  geom_cladelab(
    node = clade1,
    label = "Group A",
    align = TRUE,
    color = cross_colors[1],
    offset = 1,
    barsize = 2,
    fontsize = 4
  ) +
  geom_cladelab(
    node = clade2,
    label = "Group B",
    align = TRUE,
    color = cross_colors[3],
    offset = 1,
    barsize = 2,
    fontsize = 4
  ) +
  geom_tiplab(size = 3)
```
```{=html}
<!-- 
slide tree10a
-->
```

## Clade Labels Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center

ggtree(tree) +
  geom_hilight(node = clade1, fill = cross_colors[1], alpha = 0.3) +
  geom_hilight(node = clade2, fill = cross_colors[3], alpha = 0.3) +
  geom_cladelab(
    node = clade1,
    label = "Group A",
    align = TRUE,
    color = cross_colors[1],
    offset = 1,
    barsize = 2,
    fontsize = 4,
    fontface = "bold"
  ) +
  geom_cladelab(
    node = clade2,
    label = "Group B",
    align = TRUE,
    color = cross_colors[3],
    offset = 1,
    barsize = 2,
    fontsize = 4,
    fontface = "bold"
  ) +
  geom_tiplab(size = 3) +
  labs(title = "Tree with Clade Labels") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 16))
```
```{=html}
<!-- 
slide tree11
-->
```

## Circular Tree with Highlights

**Combine circular layout with clades:**
```{r}
#| echo: true
#| eval: false

ggtree(tree, layout = "circular") +
  geom_hilight(node = clade1, 
               fill = cross_colors[1], 
               alpha = 0.4, 
               extend = 0.3) +
  geom_hilight(node = clade2, 
               fill = cross_colors[3], 
               alpha = 0.4, 
               extend = 0.3) +
  geom_tiplab(size = 2.5, offset = 0.5) +
  labs(title = "Circular Tree with Highlighted Clades")
```

**Perfect for:**
- Large trees
- Presentation figures
- Showing major groups
```{=html}
<!-- 
slide tree11a
-->
```

## Circular Highlighted Tree Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 10
#| fig-align: center

ggtree(tree, layout = "circular") +
  geom_hilight(node = clade1, fill = cross_colors[1], alpha = 0.4, extend = 0.3) +
  geom_hilight(node = clade2, fill = cross_colors[3], alpha = 0.4, extend = 0.3) +
  geom_tiplab(size = 2.5, offset = 0.5) +
  labs(title = "Circular Tree with Clades",
       subtitle = "Compact visualization for large phylogenies") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
        plot.subtitle = element_text(hjust = 0.5, size = 12))
```
```{=html}
<!-- 
slide tree12
-->
```

## Branch Colors

**Color branches by trait or group:**
```{r}
#| echo: true
#| eval: false

# Assign groups to tips
tip_groups <- data.frame(
  label = tree$tip.label,
  group = sample(c("TypeA", "TypeB", "TypeC"), 
                 length(tree$tip.label), 
                 replace = TRUE)
)

# Plot with colored branches
ggtree(tree) %<+% tip_groups +
  geom_tiplab(aes(color = group), size = 3) +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(title = "Tree with Colored Tip Labels",
       color = "Type")
```

**`%<+%` operator:** Joins tree with metadata
```{=html}
<!-- 
slide tree12a
-->
```

## Colored Branches Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center

set.seed(456)
tip_groups <- data.frame(
  label = tree$tip.label,
  group = sample(c("TypeA", "TypeB", "TypeC"), 
                 length(tree$tip.label), 
                 replace = TRUE)
)

ggtree(tree) %<+% tip_groups +
  geom_tiplab(aes(color = group), size = 3, fontface = "bold") +
  scale_color_manual(values = cross_colors[c(1, 3, 5)]) +
  labs(title = "Phylogenetic Tree with Metadata",
       subtitle = "Tip colors show classification",
       color = "Type") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "right"
  )
```
```{=html}
<!-- 
slide tree13
-->
```

## Adding Node Labels

**Show support values:**
```{r}
#| echo: true
#| eval: false

# Add bootstrap values to nodes
tree$node.label <- sample(60:100, tree$Nnode, replace = TRUE)

ggtree(tree) +
  geom_tiplab(size = 3) +
  geom_nodelab(aes(label = label), 
               size = 2.5,
               color = "red",
               nudge_x = -0.5,
               nudge_y = 0.5) +
  labs(title = "Tree with Node Support Values")
```

**Common node labels:**
- Bootstrap values (0-100)
- Posterior probabilities (0-1)
- Node ages
```{=html}
<!-- 
slide tree13a
-->
```

## Node Labels Output
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
#| fig-align: center

tree$node.label <- sample(60:100, tree$Nnode, replace = TRUE)

ggtree(tree) +
  geom_tiplab(size = 3) +
  geom_nodelab(aes(label = label), 
               size = 2.5,
               color = "#d1422f",
               nudge_x = -0.5,
               nudge_y = 0.5) +
  labs(title = "Tree with Bootstrap Support Values",
       subtitle = "Red numbers show node confidence (60-100%)") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
        plot.subtitle = element_text(hjust = 0.5, size = 12))
```
```{=html}
<!-- 
slide tree14
-->
```

## Radial Tree with All Features

**Combine multiple elements:**
```{r}
#| echo: true
#| eval: false

ggtree(tree, layout = "radial") +
  geom_hilight(node = clade1, fill = cross_colors[1], alpha = 0.3) +
  geom_hilight(node = clade2, fill = cross_colors[3], alpha = 0.3) +
  geom_tiplab(size = 2, offset = 0.3) +
  geom_cladelab(
    node = clade1,
    label = "Clade 1",
    color = cross_colors[1],
    offset = 1.5,
    fontsize = 3.5,
    fontface = "bold"
  ) +
  geom_cladelab(
    node = clade2,
    label = "Clade 2",
    color = cross_colors[3],
    offset = 1.5,
    fontsize = 3.5,
    fontface = "bold"
  ) +
  labs(title = "Enhanced Radial Tree") +
  theme(plot.margin = margin(50, 50, 50, 50))
```
```{=html}
<!-- 
slide tree14a
-->
```

## Enhanced Radial Tree Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 11
#| fig-align: center

ggtree(tree, layout = "radial") +
  geom_hilight(node = clade1, fill = cross_colors[1], alpha = 0.3, extend = 0.5) +
  geom_hilight(node = clade2, fill = cross_colors[3], alpha = 0.3, extend = 0.5) +
  geom_tiplab(size = 2.5, offset = 0.5) +
  geom_cladelab(
    node = clade1,
    label = "Clade 1",
    align = TRUE,
    color = cross_colors[1],
    offset = 1.5,
    barsize = 1.5,
    fontsize = 3.5,
    fontface = "bold"
  ) +
  geom_cladelab(
    node = clade2,
    label = "Clade 2",
    align = TRUE,
    color = cross_colors[3],
    offset = 1.5,
    barsize = 1.5,
    fontsize = 3.5,
    fontface = "bold"
  ) +
  labs(title = "Publication-Quality Radial Tree",
       subtitle = "Combining highlights, labels, and custom styling") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.margin = margin(50, 50, 50, 50)
  )
```
```{=html}
<!-- 
slide tree15
-->
```

## Exercise 1: Create Your Phylogeny

::: {.question}
**Build a tree from scratch:**
```{r}
#| echo: true
#| eval: false

library(ggtree)
library(ape)

# Generate tree with 40 tips
set.seed(999)
my_tree <- rtree(40)

# Your tasks:
# 1. Create rectangular tree with tip labels
# 2. Identify 3 interesting clades (look for node numbers)
# 3. Highlight each clade with different Cross colors
# 4. Add clade labels ("Group A", "Group B", "Group C")
# 5. Add title and increase tip label size

# Hint: To find node numbers, use:
# ggtree(my_tree) + geom_text(aes(label = node), size = 2)

# Bonus: Create circular version with same highlights
```
:::
```{=html}
<!-- 
slide tree16
-->
```

## Exercise 2: Species Tree with Traits

::: {.question}
**Visualize tree with metadata:**
```{r}
#| echo: true
#| eval: false

# Create species tree
set.seed(777)
species_tree <- rtree(25)

# Add trait data
species_data <- data.frame(
  label = species_tree$tip.label,
  habitat = sample(c("Aquatic", "Terrestrial", "Aerial"), 
                   25, replace = TRUE),
  body_size = rnorm(25, 50, 15)
)

# Your tasks:
# 1. Join tree with species_data using %<+%
# 2. Color tip labels by habitat
# 3. Size tip labels by body_size
# 4. Use fan layout
# 5. Add legend
# 6. Use Cross color palette for habitats

# Advanced:
# 7. Add heatmap track showing body_size next to tips
# 8. Add bootstrap values to nodes (random 70-100)

# Hint for heatmap:
# library(ggtreeExtra)
# geom_fruit(data = species_data, geom = geom_tile, 
#            mapping = aes(y = label, fill = body_size))
```
:::
```{=html}
<!-- 
slide tree17
-->
```

## Finding Node Numbers

**How to identify nodes for highlighting:**
```{r}
#| echo: true
#| eval: false

# Method 1: Show all node numbers
ggtree(tree) + 
  geom_text(aes(label = node), size = 2, color = "red")

# Method 2: Interactive selection
library(ggtreeExtra)
identify_tree <- ggtree(tree) + geom_tiplab()
# Click on branches to see node numbers

# Method 3: Find MRCA (Most Recent Common Ancestor)
MRCA(tree, c("t1", "t5", "t8"))  # Returns node number

# Method 4: Get offspring of a node
offspring(tree, node = 35)  # All tips descending from node 35
```
```{=html}
<!-- 
slide tree18
-->
```

## Advanced: Time-Scaled Trees

**Show evolutionary time:**
```{r}
#| echo: true
#| eval: false

# Tree with branch lengths = time
ggtree(tree) +
  geom_tiplab(size = 3) +
  theme_tree2() +  # Add time axis
  labs(title = "Time-Calibrated Phylogeny",
       x = "Time (millions of years ago)")

# Add geological time scale
library(deeptime)
ggtree(tree) +
  coord_geo(xlim = c(0, max(node.depth.edgelength(tree)))) +
  geom_tiplab(size = 3)
```

**Use when:**
- Branch lengths = time
- Want to show divergence dates
- Comparing rates of evolution
```{=html}
<!-- 
slide tree19
-->
```

## Exporting High-Quality Trees

**Save for publication:**
```{r}
#| echo: true
#| eval: false

# Create your final tree
final_tree <- ggtree(tree, layout = "circular") +
  geom_hilight(node = clade1, fill = cross_colors[1], alpha = 0.3) +
  geom_hilight(node = clade2, fill = cross_colors[3], alpha = 0.3) +
  geom_tiplab(size = 2.5, offset = 0.5) +
  labs(title = "My Phylogenetic Tree")

# Save as PNG (high resolution)
ggsave("tree_figure.png", 
       plot = final_tree, 
       width = 10, 
       height = 10, 
       dpi = 300)

# Save as PDF (vector, scalable)
ggsave("tree_figure.pdf", 
       plot = final_tree, 
       width = 10, 
       height = 10)

# Save as SVG (for editing in Illustrator/Inkscape)
ggsave("tree_figure.svg", 
       plot = final_tree, 
       width = 10, 
       height = 10)
```
```{=html}
<!-- 
slide tree20
-->
```

## Summary: Tree Visualization

**Key components:**

✅ **Layouts**: rectangular, circular, radial, fan  
✅ **Highlighting**: `geom_hilight()` for clades  
✅ **Labels**: `geom_tiplab()`, `geom_cladelab()`  
✅ **Metadata**: Join with `%<+%` operator  
✅ **Customization**: Full ggplot2 control  
✅ **Export**: High-resolution for publication

**ggtree advantages:**

- ggplot2 syntax (familiar!)
- Highly customizable
- Integrates with tidyverse
- Publication-ready output

**Next steps:** Explore ggtreeExtra for advanced annotations!


```{=html}
<!-- 
slide tree21
-->
```

# Advanced: ggtreeExtra {background-color="#28A87D"}
```{=html}
<!-- 
slide tree22
-->
```

## What is ggtreeExtra?

:::::: columns
::: {.column width="50%"}
**Purpose:**

Add data layers around phylogenetic trees

::: fragment
**What you can add:**

- Heatmaps (gene expression, traits)
- Bar plots (abundances, counts)
- Dot plots (presence/absence)
- Point annotations (categories)
- Any data visualization!
:::
:::

::: {.column width="50%"}
::: fragment
**Why use it?**

- **Integrate multi-omics data**
- Show trait evolution
- Compare gene expression
- Display metadata visually
- Publication-quality figures

**We'll use:**

- Manual annotation (most stable)
- Direct geom placement
- Works with all ggplot2 versions
:::
:::
::::::
```{=html}
<!-- 
slide tree23
-->
```

## Setup: Tree + Data

**Prepare tree and metadata:**
```{r}
#| echo: true
#| eval: false

library(ggtree)
library(ggplot2)

# Create tree
set.seed(123)
tree <- rtree(30)

# Create trait data
trait_data <- data.frame(
  label = tree$tip.label,
  expression = rnorm(30, 50, 15),
  abundance = rpois(30, 20),
  habitat = sample(c("Forest", "Grassland", "Desert"), 
                   30, replace = TRUE),
  body_size = rnorm(30, 100, 25)
)

# Join tree with data using %<+%
tree_plot <- ggtree(tree) %<+% trait_data
```

**Key:** Use `%<+%` to join tree with metadata
```{=html}
<!-- 
slide tree24
-->
```

## Method: Getting Tip Positions

**Extract y-coordinates for annotations:**
```{r}
#| echo: true
#| eval: false

# Get tree structure data
tree_data <- fortify(tree)

# Get only tip data
tip_data <- tree_data[tree_data$isTip, ]

# Merge with trait data
tip_data <- merge(tip_data, trait_data, by = "label")

# Now tip_data has:
# - y: vertical position of each tip
# - x: horizontal position (branch length)
# - label: tip name
# - All trait columns

head(tip_data[, c("label", "x", "y", "expression", "habitat")])
```

**This gives us coordinates to place annotations!**
```{=html}
<!-- 
slide tree25
-->
```

## Example 1: Heatmap Track - Code

**Add heatmap using geom_tile:**
```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "1-4|6-8|10-15|17-19"

# Get tree and tip positions
tree_data <- fortify(tree)
tip_data <- merge(tree_data[tree_data$isTip, ], 
                  trait_data, by = "label")

# Find max x position for placement
max_x <- max(tree_data$x)

# Create plot
ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 1.5) +
  geom_tile(data = tip_data,
           aes(x = max_x + 2, y = y, fill = expression),
           width = 1, height = 0.8) +
  
  scale_fill_gradientn(colors = cross_colors) +
  xlim(NA, max_x + 4) +  # Extend x-axis for heatmap
  labs(title = "Tree with Expression Heatmap")
```
```{=html}
<!-- 
slide tree25a
-->
```

## Example 1: Heatmap Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 11
#| fig-align: center
#| warning: false
#| message: false

library(ggtree)

set.seed(123)
tree <- rtree(30)

trait_data <- data.frame(
  label = tree$tip.label,
  expression = rnorm(30, 50, 15)
)

tree_data <- fortify(tree)
tip_data <- merge(tree_data[tree_data$isTip, ], trait_data, by = "label")
max_x <- max(tree_data$x)

cross_colors <- met.brewer("Cross", 5)

ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 1.5) +
  geom_tile(data = tip_data,
           aes(x = max_x + 2, y = y, fill = expression),
           width = 1, height = 0.8) +
  scale_fill_gradientn(
    colors = cross_colors,
    name = "Expression\nLevel"
  ) +
  xlim(NA, max_x + 4) +
  labs(title = "Phylogeny with Gene Expression Heatmap",
       subtitle = "Color intensity shows expression level") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "right"
  )
```
```{=html}
<!-- 
slide tree26
-->
```

## Example 2: Bar Plot Track - Code

**Add bars using geom_segment:**
```{r}
#| echo: true
#| eval: false

# Add abundance data
trait_data$abundance <- rpois(30, 20)
tip_data <- merge(tree_data[tree_data$isTip, ], 
                  trait_data, by = "label")

ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 2) +
  geom_segment(
    data = tip_data,
    aes(x = max_x + 1, 
        xend = max_x + 1 + abundance/5,  # Scale bars
        y = y, 
        yend = y),
    color = cross_colors[3],
    linewidth = 2
  ) +
  xlim(NA, max_x + 6)
```

**Key:** Use `geom_segment` for horizontal bars
```{=html}
<!-- 
slide tree26a
-->
```

## Example 2: Bar Plot Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 11
#| fig-align: center
#| warning: false
#| message: false

trait_data$abundance <- rpois(30, 20)
tip_data <- merge(tree_data[tree_data$isTip, ], trait_data, by = "label")

ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 2) +
  geom_segment(data = tip_data,
              aes(x = max_x + 1, 
                  xend = max_x + 1 + abundance/5,
                  y = y, yend = y),
              color = cross_colors[3],
              linewidth = 2) +
  xlim(NA, max_x + 6) +
  labs(title = "Phylogeny with Abundance Bar Plots",
       subtitle = "Bar length shows relative abundance") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  )
```
```{=html}
<!-- 
slide tree27
-->
```

## Example 3: Point Track - Code

**Show categorical data with points:**
```{r}
#| echo: true
#| eval: false

# Add habitat and body size
trait_data$habitat <- sample(c("Forest", "Grassland", "Desert"), 
                             30, replace = TRUE)
trait_data$body_size <- rnorm(30, 100, 25)
tip_data <- merge(tree_data[tree_data$isTip, ], 
                  trait_data, by = "label")

ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 1.5) +
  geom_point(
    data = tip_data,
    aes(x = max_x + 2, y = y, 
        color = habitat, 
        size = body_size),
    alpha = 0.8
  ) +
  scale_color_manual(values = cross_colors[c(1,3,5)]) +
  scale_size_continuous(range = c(3, 10)) +
  xlim(NA, max_x + 4)
```
```{=html}
<!-- 
slide tree27a
-->
```

## Example 3: Point Track Output
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 11
#| fig-align: center
#| warning: false
#| message: false

set.seed(456)
trait_data$habitat <- sample(c("Forest", "Grassland", "Desert"), 
                             30, replace = TRUE)
trait_data$body_size <- rnorm(30, 100, 25)
tip_data <- merge(tree_data[tree_data$isTip, ], trait_data, by = "label")

ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 1.5) +
  geom_point(data = tip_data,
            aes(x = max_x + 2, y = y, 
                color = habitat, size = body_size),
            alpha = 0.8) +
  scale_color_manual(values = cross_colors[c(1,3,5)], 
                    name = "Habitat") +
  scale_size_continuous(range = c(3, 10), 
                       name = "Body Size") +
  xlim(NA, max_x + 4) +
  labs(title = "Phylogeny with Trait Annotations",
       subtitle = "Point color = habitat, size = body size") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "right"
  )
```
```{=html}
<!-- 
slide tree28
-->
```

## Example 4: Multiple Tracks - Code

**Combine heatmap + bars + points:**
```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "1-3|5-8|10-15|17-21"

# Get all data together
tip_data <- merge(tree_data[tree_data$isTip, ], 
                  trait_data, by = "label")

# Track 1: Expression heatmap
track1_x <- max_x + 1.5
# Track 2: Abundance bars
track2_x <- max_x + 3.5
# Track 3: Habitat points
track3_x <- max_x + 6

ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2, offset = 1) +
  # Track 1: Heatmap
  geom_tile(data = tip_data,
           aes(x = track1_x, y = y, fill = expression),
           width = 0.8, height = 0.8) +
  scale_fill_gradientn(colors = cross_colors, 
                      name = "Expression") +
  # Track 2: Bars
  new_scale_fill() +  # Reset fill scale
  geom_segment(data = tip_data,
              aes(x = track2_x, 
                  xend = track2_x + abundance/5,
                  y = y, yend = y, color = habitat),
              linewidth = 2) +
  scale_color_manual(values = cross_colors[c(1,3,5)], 
                    name = "Habitat") +
  xlim(NA, max_x + 8)
```

**Important:** Use `library(ggnewscale)` and `new_scale_fill()` for multiple color scales
```{=html}
<!-- 
slide tree28a
-->
```

## Example 4: Multiple Tracks Output
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 12
#| fig-align: center
#| warning: false
#| message: false

library(ggnewscale)

tip_data <- merge(tree_data[tree_data$isTip, ], trait_data, by = "label")
track1_x <- max_x + 1.5
track2_x <- max_x + 3.5

ggtree(tree, layout = "circular") +
  geom_tiplab(size = 2.5, offset = 1) +
  # Track 1: Expression heatmap
  geom_tile(data = tip_data,
           aes(x = track1_x, y = y, fill = expression),
           width = 0.8, height = 0.8) +
  scale_fill_gradientn(colors = cross_colors, 
                      name = "Expression") +
  # Track 2: Abundance bars colored by habitat
  new_scale_fill() +
  geom_segment(data = tip_data,
              aes(x = track2_x, 
                  xend = track2_x + abundance/5,
                  y = y, yend = y, color = habitat),
              linewidth = 2) +
  scale_color_manual(values = cross_colors[c(1,3,5)], 
                    name = "Habitat") +
  xlim(NA, max_x + 8) +
  labs(title = "Multi-layer Phylogenetic Visualization",
       subtitle = "Expression heatmap + abundance bars colored by habitat") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "right"
  )
```
```{=html}
<!-- 
slide tree29
-->
```

## Example 5: Rectangular Layout with Tracks

**Same approach works for rectangular trees:**
```{r}
#| echo: true
#| eval: false

# Rectangular tree
tree_data_rect <- fortify(tree)
tip_data_rect <- merge(tree_data_rect[tree_data_rect$isTip, ], 
                       trait_data, by = "label")
max_x_rect <- max(tree_data_rect$x)

ggtree(tree) +
  geom_tiplab(size = 3, offset = 0.5) +
  # Heatmap track
  geom_tile(data = tip_data_rect,
           aes(x = max_x_rect + 1, y = y, fill = expression),
           width = 0.5, height = 0.8) +
  scale_fill_gradientn(colors = cross_colors) +
  # Bar track
  geom_segment(data = tip_data_rect,
              aes(x = max_x_rect + 2, 
                  xend = max_x_rect + 2 + abundance/10,
                  y = y, yend = y),
              color = cross_colors[3], linewidth = 1.5) +
  xlim(NA, max_x_rect + 6)
```
```{=html}
<!-- 
slide tree29a
-->
```

## Rectangular with Tracks Output
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 9
#| fig-align: center
#| warning: false
#| message: false

tree_data_rect <- fortify(tree)
tip_data_rect <- merge(tree_data_rect[tree_data_rect$isTip, ], 
                       trait_data, by = "label")
max_x_rect <- max(tree_data_rect$x)

ggtree(tree) +
  geom_tiplab(size = 3, offset = 0.5) +
  # Heatmap
  geom_tile(data = tip_data_rect,
           aes(x = max_x_rect + 1, y = y, fill = expression),
           width = 0.5, height = 0.8) +
  scale_fill_gradientn(colors = cross_colors, 
                      name = "Expression") +
  # Bars
  geom_segment(data = tip_data_rect,
              aes(x = max_x_rect + 2, 
                  xend = max_x_rect + 2 + abundance/10,
                  y = y, yend = y),
              color = cross_colors[3], linewidth = 1.5) +
  xlim(NA, max_x_rect + 6) +
  labs(title = "Rectangular Tree with Data Tracks",
       subtitle = "Expression heatmap + abundance bars") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "right"
  )
```
```{=html}
<!-- 
slide tree30
-->
```

## Exercise 3: Multi-Omics Tree

::: {.question}
**Create comprehensive visualization:**
```{r}
#| echo: true
#| eval: false

# Prepare multi-omics data
set.seed(2024)
omics_tree <- rtree(25)

omics_data <- data.frame(
  label = omics_tree$tip.label,
  # Transcriptomics
  gene1_expr = rnorm(25, 100, 30),
  gene2_expr = rnorm(25, 150, 40),
  # Proteomics
  protein_abundance = rpois(25, 50),
  # Phenotype
  phenotype = sample(c("Wild", "Mutant"), 25, replace = TRUE)
)

# Your tasks:
# 1. Get tree tip positions using fortify() and merge with omics_data
# 2. Create circular tree with tip labels
# 3. Add heatmap for gene1_expr (use geom_tile)
# 4. Add heatmap for gene2_expr next to it
# 5. Add bars for protein_abundance (use geom_segment)
# 6. Add points colored by phenotype
# 7. Space tracks appropriately (use max_x + offsets)

# Hint: Calculate positions like this:
# track1_x <- max_x + 1.5
# track2_x <- max_x + 2.5
# track3_x <- max_x + 4
# track4_x <- max_x + 6
```
:::
```{=html}
<!-- 
slide tree31
-->
```

## Exercise 4: Gene Presence/Absence

::: {.question}
**Visualize gene family patterns:**
```{r}
#| echo: true
#| eval: false

# Gene family data
set.seed(3030)
gene_tree <- rtree(20)

# Create presence/absence matrix
gene_families <- expand.grid(
  label = gene_tree$tip.label,
  family = paste0("Family_", LETTERS[1:6])
) |>
  mutate(
    presence = sample(c(0, 1), n(), replace = TRUE, 
                     prob = c(0.3, 0.7))
  )

# Your tasks:
# 1. Get tree positions and merge with gene_families
# 2. Create fan layout tree
# 3. Add 6 heatmap columns (one per gene family)
#    - Use different x positions for each family
#    - Color: 0 = white, 1 = colored
# 4. Highlight 2 major clades
# 5. Add clade labels

# Hint: For binary heatmap
# scale_fill_gradient(low = "white", high = cross_colors[1])

# Advanced: Calculate how many families each species has
# Add this as bar plot track
```
:::
```{=html}
<!-- 
slide tree32
-->
```

## Key Concepts: Manual Annotation

**What we learned:**
```{r}
#| echo: true
#| eval: false

# 1. Get tree positions
tree_data <- fortify(tree)
tip_data <- tree_data[tree_data$isTip, ]

# 2. Merge with your data
tip_data <- merge(tip_data, trait_data, by = "label")

# 3. Calculate track positions
max_x <- max(tree_data$x)
track1_x <- max_x + 2
track2_x <- max_x + 4

# 4. Add annotations
+ geom_tile(data = tip_data, 
           aes(x = track1_x, y = y, fill = value))
+ geom_segment(data = tip_data,
              aes(x = track2_x, xend = track2_x + value,
                  y = y, yend = y))
+ geom_point(data = tip_data,
            aes(x = track3_x, y = y, color = category))

# 5. Extend x-axis
+ xlim(NA, max_x + 8)
```
```{=html}
<!-- 
slide tree33
-->
```

## Best Practices

**Design guidelines:**

::: fragment
**Spacing:**

- Leave 1-2 units between tree and first track
- Leave 0.5-1 unit between tracks
- Use `xlim(NA, max_x + buffer)` to extend space
:::

::: fragment
**Multiple color scales:**

- Load `library(ggnewscale)`
- Call `new_scale_fill()` before each new scale
- Each track can have different colors
:::

::: fragment
**Track types:**

- **Heatmaps**: `geom_tile()` with continuous data
- **Bars**: `geom_segment()` for horizontal bars
- **Points**: `geom_point()` for categorical data
- **Text**: `geom_text()` for labels
:::
```{=html}
<!-- 
slide tree34
-->
```

## Complete Example: Publication Figure
```{r}
#| echo: true
#| eval: false

library(ggnewscale)

# Get data
tip_data <- merge(tree_data[tree_data$isTip, ], trait_data, by = "label")

# Calculate positions
max_x <- max(tree_data$x)

publication_tree <- ggtree(tree, layout = "circular") +
  # Tree styling
  geom_tiplab(size = 2.5, offset = 1) +
  geom_hilight(node = 35, fill = cross_colors[1], alpha = 0.2) +
  geom_hilight(node = 45, fill = cross_colors[3], alpha = 0.2) +
  
  # Track 1: Expression heatmap
  geom_tile(data = tip_data,
           aes(x = max_x + 2, y = y, fill = expression),
           width = 0.8, height = 0.8) +
  scale_fill_gradientn(colors = cross_colors, name = "Expression") +
  new_scale_fill() +
  
  # Track 2: Abundance bars by habitat
  geom_segment(data = tip_data,
              aes(x = max_x + 4, xend = max_x + 4 + abundance/5,
                  y = y, yend = y, color = habitat),
              linewidth = 2) +
  scale_color_manual(values = cross_colors[c(1,3,5)], name = "Habitat") +
  
  # Styling
  xlim(NA, max_x + 10) +
  labs(title = "Integrated Phylogenomic Analysis") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 16))

# Save
ggsave("phylo_figure.pdf", publication_tree, 
       width = 12, height = 12)
```
```{=html}
<!-- 
slide tree35
-->
```

## Summary: Adding Data to Trees

**Key steps:**

✅ **Get positions**: `fortify(tree)` → extract tips  
✅ **Merge data**: Join tree positions with trait data  
✅ **Calculate x positions**: `max_x + offset` for each track  
✅ **Add layers**: Use standard ggplot2 geoms  
✅ **Multiple scales**: `new_scale_fill()` when needed  
✅ **Extend axis**: `xlim(NA, max_x + buffer)`

**Advantages of manual approach:**

- Works with all package versions
- Full ggplot2 control
- Easy to debug
- Flexible positioning
- Publication-ready

**You now have complete control over tree visualization!** 🌳

```{=html}
<!-- 
slide circular1
-->
```

# Circular genomic plots {background-color="#28A87D"}
```{=html}
<!-- 
slide circular2
-->
```

## Why Circular Plots?

:::::: columns
::: {.column width="50%"}
**Purpose:**

Visualize whole genomes in a compact, intuitive format

::: fragment
**What you can show:**

- Chromosome structure
- Gene density
- Expression levels
- Structural variants
- Synteny between regions
- Coverage tracks
:::
:::

::: {.column width="50%"}
::: fragment
**Advantages:**

✅ Compact for large genomes  
✅ Shows all chromosomes simultaneously  
✅ Easy to spot patterns  
✅ Publication favorite  
✅ Multiple data layers

**Common uses:**

- Genome-wide associations
- Structural rearrangements
- Multi-omics integration
- Comparative genomics
:::
:::
::::::
```{=html}
<!-- 
slide circular3
-->
```

## The circlize Package

:::::: columns
::: {.column width="50%"}
**Key concepts:**

::: fragment
**Sectors** = Chromosomes

Each chromosome is a sector (slice of the circle)
:::

::: fragment
**Tracks** = Data layers

Add multiple tracks moving outward from center
:::

::: fragment
**Links** = Connections

Lines connecting regions (structural variants, synteny)
:::
:::

::: {.column width="50%"}
::: fragment
**Basic workflow:**

1. **Initialize**: Set up chromosomes
2. **Add tracks**: Gene density, heatmaps, etc.
3. **Add links**: Connect regions
4. **Clear**: Reset for next plot

**Important:** Always call `circos.clear()` after each plot!
:::
:::
::::::
```{=html}
<!-- 
slide circular4
-->
```

## Step 1: Chromosome Data

**Prepare chromosome lengths:**
```{r}
#| echo: true
#| eval: false

library(circlize)

# Chromosome data: name, start, end
chr_data <- data.frame(
  chr = c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"),
  start = c(1, 1, 1, 1, 1),
  end = c(30427671, 19698289, 23459830, 18585056, 26975502)
)

# Preview
chr_data
#    chr start      end
# 1 Chr1     1 30427671
# 2 Chr2     1 19698289
# 3 Chr3     1 23459830
# 4 Chr4     1 18585056
# 5 Chr5     1 26975502
```

**Data structure:** Three columns: chromosome name, start position, end position
```{=html}
<!-- 
slide circular5
-->
```

## Step 2: Initialize Circular Plot

**Create the basic circle:**
```{r}
#| echo: true
#| eval: false

# Initialize with chromosome data
circos.genomicInitialize(chr_data)

# That's it! You now have a circular plot with 5 sectors
```

**What this does:**

- Creates circular layout
- Divides circle into sectors (one per chromosome)
- Sets coordinate system for each sector
- Adds chromosome labels by default
```{=html}
<!-- 
slide circular5a
-->
```

## Basic Circular Plot Output
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 8
#| fig-align: center

library(circlize)

chr_data <- data.frame(
  chr = c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"),
  start = c(1, 1, 1, 1, 1),
  end = c(30427671, 19698289, 23459830, 18585056, 26975502)
)

circos.genomicInitialize(chr_data)
title("Basic Circular Genome Layout")
circos.clear()
```

**Note:** Five sectors representing five chromosomes
```{=html}
<!-- 
slide circular6
-->
```

## Step 3: Adding Links

**Connect two genomic regions:**
```{r}
#| echo: true
#| eval: false

circos.genomicInitialize(chr_data)

# Add a link between Chr4 and Chr5
circos.link(
  sector.index1 = "Chr4", 
  point1 = c(9000000, 12000000),    # Region on Chr4
  sector.index2 = "Chr5", 
  point2 = c(12000000, 15000000),   # Region on Chr5
  col = "#d1422f"                   # Link color
)

circos.clear()  # Always clear after!
```

**Use for:**
- Structural variants (translocations)
- Duplications
- Syntenic regions
```{=html}
<!-- 
slide circular6a
-->
```

## Single Link Output
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 8
#| fig-align: center

circos.genomicInitialize(chr_data)
circos.link("Chr4", c(9000000, 12000000), 
           "Chr5", c(12000000, 15000000), 
           col = "#d1422f")
title("Circular Plot with Structural Variant Link")
circos.clear()
```

**Interpretation:** Red ribbon shows connection between Chr4 and Chr5 regions
```{=html}
<!-- 
slide circular7
-->
```

## Step 4: Multiple Links

**Add many links at once:**
```{r}
#| echo: true
#| eval: false

# Source regions (BED format)
source_links <- data.frame(
  chr = c("Chr1", "Chr2", "Chr3"),
  start = c(5000000, 3000000, 8000000),
  end = c(7000000, 5000000, 10000000)
)

# Target regions
target_links <- data.frame(
  chr = c("Chr4", "Chr5", "Chr2"),
  start = c(2000000, 8000000, 12000000),
  end = c(4000000, 10000000, 14000000)
)

circos.genomicInitialize(chr_data)
circos.genomicLink(source_links, target_links, 
                  col = "#f4ab5c")
circos.clear()
```
```{=html}
<!-- 
slide circular7a
-->
```

## Multiple Links Output
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 8
#| fig-align: center

source_links <- data.frame(
  chr = c("Chr1", "Chr2", "Chr3"),
  start = c(5000000, 3000000, 8000000),
  end = c(7000000, 5000000, 10000000)
)

target_links <- data.frame(
  chr = c("Chr4", "Chr5", "Chr2"),
  start = c(2000000, 8000000, 12000000),
  end = c(4000000, 10000000, 14000000)
)

circos.genomicInitialize(chr_data)
circos.genomicLink(source_links, target_links, col = "#f4ab5c")
title("Multiple Genomic Links")
circos.clear()
```

**Use case:** Show syntenic blocks or duplicated regions
```{=html}
<!-- 
slide circular8
-->
```

## Step 5: Gene Density Track

**Show where genes are located:**
```{r}
#| echo: true
#| eval: false

# Gene positions (BED format)
gene_positions <- data.frame(
  chr = c(rep("Chr1", 30), rep("Chr2", 25), 
          rep("Chr3", 20), rep("Chr4", 18), rep("Chr5", 22))
)

# Generate random positions for demo
set.seed(123)
gene_positions$start <- sample(1:29000000, 115, replace = TRUE)
gene_positions$end <- gene_positions$start + 
  sample(500:5000, 115, replace = TRUE)

circos.genomicInitialize(chr_data)
circos.genomicDensity(
  gene_positions, 
  window.size = 1e6,        # 1 Mb windows
  col = "#3d5a80",          # Color
  track.height = 0.15       # Height of track
)
circos.clear()
```
```{=html}
<!-- 
slide circular8a
-->
```

## Gene Density Output
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 8
#| fig-align: center

gene_positions <- data.frame(
  chr = c(rep("Chr1", 30), rep("Chr2", 25), 
          rep("Chr3", 20), rep("Chr4", 18), rep("Chr5", 22))
)

set.seed(123)
gene_positions$start <- sample(1:29000000, 115, replace = TRUE)
gene_positions$end <- gene_positions$start + sample(500:5000, 115, replace = TRUE)

circos.genomicInitialize(chr_data)
circos.genomicDensity(gene_positions, 
                     window.size = 1e6, 
                     col = "#3d5a80", 
                     track.height = 0.15)
title("Gene Density Track")
circos.clear()
```

**Interpretation:** Taller peaks = more genes in that region
```{=html}
<!-- 
slide circular9
-->
```

## Step 6: Heatmap Track

**Show quantitative data (expression, methylation, etc.):**
```{r}
#| echo: true
#| eval: false

library(circlize)

# Expression data
heatmap_data <- data.frame(
  chr = rep(chr_data$chr, each = 20)
)

set.seed(456)
heatmap_data$start <- sample(1:25000000, 100, replace = TRUE)
heatmap_data$end <- heatmap_data$start + 500000
heatmap_data$value <- rnorm(100, 12, 2)

# Color function
col_fun <- colorRamp2(
  c(10, 12, 15), 
  c(cross_colors[1], "white", cross_colors[3])
)

circos.genomicInitialize(chr_data)
circos.genomicHeatmap(
  heatmap_data, 
  col = col_fun,
  side = "outside",    # Place outside
  border = "white"
)
circos.clear()
```
```{=html}
<!-- 
slide circular9a
-->
```

## Heatmap Track Output
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 8
#| fig-align: center

heatmap_data <- data.frame(
  chr = rep(chr_data$chr, each = 20)
)

set.seed(456)
heatmap_data$start <- sample(1:25000000, 100, replace = TRUE)
heatmap_data$end <- heatmap_data$start + 500000
heatmap_data$value <- rnorm(100, 12, 2)

cross_colors <- met.brewer("Cross", 5)
col_fun <- colorRamp2(c(10, 12, 15), 
                     c(cross_colors[1], "white", cross_colors[3]))

circos.genomicInitialize(chr_data)
circos.genomicHeatmap(heatmap_data, col = col_fun, 
                     side = "outside", border = "white")
title("Gene Expression Heatmap")
circos.clear()
```

**Color:** Low expression (red) → High expression (teal)
```{=html}
<!-- 
slide circular10
-->
```

## Step 7: Combining Multiple Tracks

**Layer different data types:**
```{r}
#| echo: true
#| eval: false

circos.clear()
circos.genomicInitialize(chr_data)

# Track 1: Heatmap (outermost)
circos.genomicHeatmap(heatmap_data, col = col_fun, 
                     side = "outside", border = "white")

# Track 2: Gene density
circos.genomicDensity(gene_positions, 
                     window.size = 1e6, 
                     col = "#9B2226", 
                     track.height = 0.1)

# Track 3: Links (innermost)
circos.link("Chr4", c(9000000, 12000000), 
           "Chr5", c(12000000, 15000000), 
           col = "#E28900")

circos.genomicLink(source_links, target_links, 
                  col = "#E9D8A6")

circos.clear()
```

**Order matters:** Add tracks from outside to inside
```{=html}
<!-- 
slide circular10a
-->
```

## Combined Tracks Output
```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 9
#| fig-align: center

circos.genomicInitialize(chr_data)

# Heatmap
circos.genomicHeatmap(heatmap_data, col = col_fun, 
                     side = "outside", border = "white")

# Density
circos.genomicDensity(gene_positions, window.size = 1e6, 
                     col = "#9B2226", track.height = 0.1)

# Links
circos.link("Chr4", c(9000000, 12000000), 
           "Chr5", c(12000000, 15000000), 
           col = "#E28900")
circos.genomicLink(source_links, target_links, col = "#E9D8A6")

title("Multi-layer Circular Genome Plot")
circos.clear()
```

**All together:** Expression + density + structural variants
```{=html}
<!-- 
slide circular11
-->
```

## Step 8: Custom Track with Lines

**Show coverage or signal tracks:**
```{r}
#| echo: true
#| eval: false

# Generate coverage data
set.seed(789)
coverage_data <- data.frame(
  chr = rep(chr_data$chr, each = 50)
)
coverage_data$pos <- rep(seq(1, 30000000, length.out = 50), 5)
coverage_data$coverage <- abs(rnorm(250, 50, 20))

circos.genomicInitialize(chr_data)

# Add custom track with lines
circos.genomicTrack(
  coverage_data,
  panel.fun = function(region, value, ...) {
    circos.genomicLines(region, value, 
                       col = cross_colors[3], 
                       lwd = 2)
  }
)

circos.clear()
```

**`panel.fun`:** Custom function to draw in each sector
```{=html}
<!-- 
slide circular11a
-->
```

## Coverage Track Output
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 8
#| fig-align: center

set.seed(789)
coverage_data <- data.frame(
  chr = rep(chr_data$chr, each = 50)
)
coverage_data$pos <- rep(seq(1, 30000000, length.out = 50), 5)
coverage_data$value <- abs(rnorm(250, 50, 20))
coverage_data <- coverage_data[, c("chr", "pos", "pos", "value")]
names(coverage_data) <- c("chr", "start", "end", "value")

circos.genomicInitialize(chr_data)

circos.genomicTrack(
  coverage_data,
  panel.fun = function(region, value, ...) {
    circos.genomicLines(region, value, col = cross_colors[3], lwd = 2)
  }
)

title("Sequencing Coverage Track")
circos.clear()
```
```{=html}
<!-- 
slide circular12
-->
```

## Exercise 1: Basic Circular Plot

::: {.question}
**Create a multi-track visualization:**
```{r}
#| echo: true
#| eval: false

library(circlize)

# Three-chromosome genome
chr_data <- data.frame(
  chr = c("Chr1", "Chr2", "Chr3"),
  start = c(1, 1, 1),
  end = c(25000000, 20000000, 18000000)
)

# Generate random gene positions
set.seed(101)
genes <- data.frame(
  chr = sample(c("Chr1", "Chr2", "Chr3"), 100, replace = TRUE)
)
genes$start <- sample(1:24000000, 100, replace = TRUE)
genes$end <- genes$start + sample(1000:5000, 100, replace = TRUE)

# Generate expression data
expr_data <- data.frame(
  chr = sample(c("Chr1", "Chr2", "Chr3"), 60, replace = TRUE)
)
expr_data$start <- sample(1:24000000, 60, replace = TRUE)
expr_data$end <- expr_data$start + 500000
expr_data$value <- rnorm(60, 50, 15)

# Your tasks:
# 1. Initialize circular plot with chr_data
# 2. Add heatmap track with expr_data (use colorRamp2)
# 3. Add gene density track
# 4. Add a link between Chr1 (5M-8M) and Chr3 (10M-13M)
# 5. Clear plot when done

# Use Cross color palette throughout
```
:::
```{=html}
<!-- 
slide circular13
-->
```

## Exercise 2: Comparative Genomics

::: {.question}
**Show synteny between two regions:**
```{r}
#| echo: true
#| eval: false

# Five chromosomes
chr_data <- data.frame(
  chr = paste0("Chr", 1:5),
  start = rep(1, 5),
  end = c(30000000, 25000000, 22000000, 20000000, 18000000)
)

# Create syntenic links (10 blocks)
set.seed(202)
synteny_source <- data.frame(
  chr = sample(paste0("Chr", 1:3), 10, replace = TRUE),
  start = sample(1:20000000, 10),
  end = sample(5000000:25000000, 10)
)

synteny_target <- data.frame(
  chr = sample(paste0("Chr", 4:5), 10, replace = TRUE),
  start = sample(1:15000000, 10),
  end = sample(3000000:17000000, 10)
)

# Your tasks:
# 1. Initialize plot
# 2. Add all syntenic links with semi-transparent color
# 3. Color links by source chromosome (Chr1=color1, Chr2=color2, Chr3=color3)
# 4. Add gene density track for all chromosomes
# 5. Add title "Syntenic Blocks Between Species"

# Bonus: Add a heatmap showing SNP density
```
:::
```{=html}
<!-- 
slide circular14
-->
```

## Common Track Types
```{r}
#| echo: false

library(gt)

track_types <- tibble(
  Track = c("circos.genomicDensity()", "circos.genomicHeatmap()", 
            "circos.genomicLines()", "circos.genomicPoints()", 
            "circos.genomicTrack()"),
  `Data Type` = c(
    "Intervals (BED)",
    "Values + positions",
    "Continuous signal",
    "Discrete points",
    "Custom (any)"
  ),
  `Best For` = c(
    "Gene/feature density",
    "Expression, methylation",
    "Coverage, signal",
    "SNPs, mutations",
    "Custom visualizations"
  ),
  Example = c(
    "Gene distribution",
    "RNA-seq heatmap",
    "ChIP-seq signal",
    "Variant positions",
    "Custom plots"
  )
)

track_types |>
  gt() |>
  cols_align(align = "left") |>
  tab_style(
    style = cell_text(weight = "bold", size = px(13)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(size = px(11)),
    locations = cells_body()
  ) |>
  tab_options(
    table.font.size = px(12),
    table.width = pct(95)
  )
```
```{=html}
<!-- 
slide circular15
-->
```

## Customization Options

**Control plot appearance:**
```{r}
#| echo: true
#| eval: false

# Set global parameters BEFORE circos.genomicInitialize()
circos.par(
  start.degree = 90,        # Rotate plot (0-360)
  gap.degree = 4,           # Space between sectors
  track.height = 0.15,      # Default track height
  cell.padding = c(0, 0, 0, 0)  # Padding
)

circos.genomicInitialize(chr_data)

# Individual track parameters
circos.genomicDensity(
  gene_positions,
  track.height = 0.2,       # Override default
  bg.border = "black",      # Border color
  bg.col = "grey95"         # Background color
)

circos.clear()
```

**Always set `circos.par()` BEFORE initializing!**
```{=html}
<!-- 
slide circular16
-->
```

## Best Practices

**Design guidelines:**

::: fragment
**1. Order of elements:**

- Initialize first
- Add heatmaps/tracks (outside → inside)
- Add links last
- Always `circos.clear()` at end
:::

::: fragment
**2. Colors:**

- Use consistent palette (Cross colors!)
- Semi-transparent links (alpha = 0.5-0.7)
- Contrasting colors for different tracks
:::

::: fragment
**3. Track heights:**

- Balance track sizes (0.1-0.2 typical)
- More important data = bigger track
- Don't overcrowd
:::
```{=html}
<!-- 
slide circular17
-->
```

## Troubleshooting

**Common issues:**

::: fragment
**Problem:** Plot doesn't clear / elements overlap

**Solution:** Always call `circos.clear()` after each plot
```{r}
#| eval: false
circos.genomicInitialize(chr_data)
# ... add tracks ...
circos.clear()  # IMPORTANT!
```
:::

::: fragment
**Problem:** Track not visible

**Solution:** Increase `track.height` parameter
```{r}
#| eval: false
circos.genomicDensity(data, track.height = 0.2)  # Bigger
```
:::

::: fragment
**Problem:** Links go to wrong positions

**Solution:** Check chromosome names match exactly
```{r}
#| eval: false
# Make sure names match
unique(chr_data$chr)
unique(link_data$chr)
```
:::
```{=html}
<!-- 
slide circular18
-->
```

## Complete Example: Publication Figure
```{r}
#| echo: true
#| eval: false

library(circlize)

# Set parameters
circos.par(start.degree = 90, gap.degree = 3)

# Initialize
circos.genomicInitialize(chr_data)

# Track 1: Expression heatmap (outside)
col_fun <- colorRamp2(c(10, 50, 90), 
                     c(cross_colors[1], "white", cross_colors[3]))
circos.genomicHeatmap(expr_data, col = col_fun, 
                     side = "outside", border = "white")

# Track 2: Gene density
circos.genomicDensity(gene_positions, 
                     window.size = 1e6,
                     col = cross_colors[5],
                     track.height = 0.15)

# Track 3: SNP positions
circos.genomicPoints(snp_data, 
                    pch = 16, cex = 0.3,
                    col = "#d1422f")

# Links: Structural variants
circos.genomicLink(sv_source, sv_target, 
                  col = "#f4ab5c80")  # 80 = alpha

# Title
title("Integrated Genome View", cex.main = 1.5)

circos.clear()
```
```{=html}
<!-- 
slide circular19
-->
```

## Key Takeaways

**Remember:**

✅ **Initialize** with `circos.genomicInitialize()`  
✅ **Add tracks** from outside to inside  
✅ **Use links** for connections  
✅ **Always clear** with `circos.clear()`  
✅ **Set parameters** before initializing

**Common workflow:**

1. Prepare chromosome data
2. Set global parameters
3. Initialize plot
4. Add tracks (heatmaps, density, etc.)
5. Add links
6. Clear plot

**Perfect for genome-wide visualization!** ⭕



```{=html}
<!--
                                        
  __ _  __ _  __ _  ___ _ __   ___  ___ 
 / _` |/ _` |/ _` |/ _ \ '_ \ / _ \/ __|
| (_| | (_| | (_| |  __/ | | |  __/\__ \
 \__, |\__, |\__, |\___|_| |_|\___||___/
  __/ | __/ | __/ |                     
 |___/ |___/ |___/                      
 
 -->
```

# Gene arrangements & synteny {background-color="#28A87D"}


```{=html}
<!-- 
slide synteny2
-->
```

## What is Synteny?

:::::: columns
::: {.column width="50%"}
**Definition:**

Synteny = Conservation of gene order and chromosomal organization between species or within genomes

::: fragment
**Why study synteny?**

- Track chromosomal rearrangements
- Validate genome assemblies
- Predict gene functions
- Identify conserved regions
:::
:::

::: {.column width="50%"}
::: fragment
**Types of synteny:**

1. **Conserved**: Same order, same orientation
2. **Inverted**: Same genes, flipped orientation
3. **Disrupted**: Rearrangements, insertions, deletions
:::
:::
::::::
```{=html}
<!-- 
slide synteny2a
-->
```

## Why Visualize Gene Arrangements?

**Common biological questions:**

:::::: columns
::: {.column width="50%"}
::: fragment
**Evolutionary studies:**
- Which genes moved?
- Where did inversions occur?
- What blocks are conserved?
:::

::: fragment
**Functional genomics:**
- Are genes in operons?
- Which genes cluster together?
- Are pathways co-located?
:::
:::

::: {.column width="50%"}
::: fragment
**Genome annotation:**
- Transfer annotations across species
- Identify mobile elements
- Find insertion/deletion events
:::

::: fragment
**Today we'll learn:**
- `gggenes` for gene-level detail
- How to show rearrangements
- How to highlight insertions (prophages!)
:::
:::
::::::
```{=html}
<!-- 
slide synteny3
-->
```

## gggenes: Basic Syntax

**Simple and integrates with ggplot2:**
```{r}
#| echo: true
#| eval: false

library(gggenes)

# Required data columns
data.frame(
  molecule = "Genome_A",    # Which genome/chromosome
  gene = "Gene1",           # Gene name
  start = 1000,             # Start position (bp)
  end = 2000,               # End position (bp)
  orientation = 1           # 1 = forward (→), -1 = reverse (←)
)

# Required aesthetics in ggplot
aes(xmin = start, xmax = end,    # Gene boundaries
    y = molecule,                 # Which genome
    forward = orientation)        # Arrow direction
```
```{=html}
<!-- 
slide synteny4
-->
```

## Example 1: Single Chromosome

**Code:**
```{r}
#| echo: true
#| eval: false

library(gggenes)

genes <- data.frame(
  molecule = "Chromosome_1",
  gene = paste0("Gene_", LETTERS[1:5]),
  start = c(1000, 3000, 5500, 8000, 10000),
  end = c(2500, 4800, 7200, 9500, 11500),
  orientation = c(1, 1, -1, 1, -1)  # Mix of forward/reverse
)

ggplot(genes, aes(xmin = start, xmax = end, y = molecule,
                  fill = gene, forward = orientation)) +
  geom_gene_arrow() +
  scale_fill_manual(values = met.brewer("Cross", 5)) +
  theme_genes()
```
```{=html}
<!-- 
slide synteny4a
-->
```

## Example 1: Output
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 3
#| fig-align: center

library(gggenes)

genes <- data.frame(
  molecule = "Chromosome_1",
  gene = paste0("Gene_", LETTERS[1:5]),
  start = c(1000, 3000, 5500, 8000, 10000),
  end = c(2500, 4800, 7200, 9500, 11500),
  orientation = c(1, 1, -1, 1, -1)
)

cross_colors <- met.brewer("Cross", 5)

ggplot(genes, aes(xmin = start, xmax = end, y = molecule,
                  fill = gene, forward = orientation)) +
  geom_gene_arrow(arrowhead_height = unit(4, "mm"),
                  arrowhead_width = unit(2, "mm")) +
  scale_fill_manual(values = cross_colors) +
  labs(title = "Gene Arrangement on Chromosome 1",
       x = "Position (bp)", y = "") +
  theme_genes() +
  theme(legend.position = "bottom")
```

**Note:** Gene C points left (←) = reverse strand
```{=html}
<!-- 
slide synteny5
-->
```

## Example 2: Comparing Species

**Code:**
```{r}
#| echo: true
#| eval: false

multi_genome <- data.frame(
  molecule = rep(c("Species_A", "Species_B", "Species_C"), each = 5),
  gene = rep(paste0("Gene_", LETTERS[1:5]), 3),
  start = c(
    c(1000, 3000, 5500, 8000, 10000),  # Species A (reference)
    c(1100, 3200, 5800, 8500, 10500),  # Species B (Gene C moved)
    c(1000, 3000, 8000, 10200, 5500)   # Species C (D-E swapped)
  ),
  end = c(
    c(2500, 4800, 7200, 9500, 11500),
    c(2600, 5000, 7500, 10200, 12000),
    c(2500, 4800, 9500, 11700, 7200)
  ),
  orientation = rep(c(1, 1, -1, 1, -1), 3)
)

ggplot(multi_genome, aes(xmin = start, xmax = end, y = molecule,
                         fill = gene, forward = orientation)) +
  geom_gene_arrow() +
  facet_wrap(~ molecule, scales = "free_y", ncol = 1)
```
```{=html}
<!-- 
slide synteny5a
-->
```

## Example 2: Output
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
#| fig-align: center

multi_genome <- data.frame(
  molecule = rep(c("Species_A", "Species_B", "Species_C"), each = 5),
  gene = rep(paste0("Gene_", LETTERS[1:5]), 3),
  start = c(
    c(1000, 3000, 5500, 8000, 10000),
    c(1100, 3200, 5800, 8500, 10500),
    c(1000, 3000, 8000, 10200, 5500)
  ),
  end = c(
    c(2500, 4800, 7200, 9500, 11500),
    c(2600, 5000, 7500, 10200, 12000),
    c(2500, 4800, 9500, 11700, 7200)
  ),
  orientation = rep(c(1, 1, -1, 1, -1), 3)
)

cross_colors <- met.brewer("Cross", 5)

ggplot(multi_genome, aes(xmin = start, xmax = end, y = molecule,
                         fill = gene, forward = orientation)) +
  geom_gene_arrow(arrowhead_height = unit(4, "mm"),
                  arrowhead_width = unit(2, "mm")) +
  facet_wrap(~ molecule, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = cross_colors) +
  labs(title = "Gene Rearrangements Across Species",
       x = "Position (bp)", y = "") +
  theme_genes() +
  theme(legend.position = "bottom")
```

**Spot the differences:** Gene C moved in Species B, Genes D-E swapped in Species C
```{=html}
<!-- 
slide synteny6
-->
```

## Example 3: Prophage Insertion

**Biological scenario:** Prophage integrates into bacterial chromosome

**Code:**
```{r}
#| echo: true
#| eval: false

prophage_data <- data.frame(
  strain = rep(c("Wildtype", "Lysogen"), each = 10),
  gene = c(
    # Wildtype: normal genes
    paste0("Gene_", 1:10),
    # Lysogen: genes + prophage inserted between Gene_4 and Gene_5
    c(paste0("Gene_", 1:4), 
      paste0("Phage_", 1:3),
      paste0("Gene_", 5:7))
  ),
  start = c(
    seq(1000, 50000, length.out = 10),  # Wildtype
    c(seq(1000, 20000, length.out = 4),  # Genes 1-4
      seq(22000, 34000, length.out = 3),  # Prophage genes
      seq(36000, 50000, length.out = 3))  # Genes 5-7
  ),
  type = c(
    rep("host", 10),
    c(rep("host", 4), rep("prophage", 3), rep("host", 3))
  )
)
prophage_data$end <- prophage_data$start + 
  ifelse(prophage_data$type == "prophage", 4000, 3000)
prophage_data$orientation <- 1
```
```{=html}
<!-- 
slide synteny6a
-->
```

## Example 3: Prophage Output
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
#| fig-align: center

prophage_data <- data.frame(
  strain = rep(c("Wildtype", "Lysogen"), c(10, 10)),
  gene = c(
    paste0("Gene_", 1:10),
    c(paste0("Gene_", 1:4), paste0("Phage_", 1:3), paste0("Gene_", 5:7))
  ),
  start = c(
    seq(1000, 50000, length.out = 10),
    c(seq(1000, 20000, length.out = 4),
      seq(22000, 34000, length.out = 3),
      seq(36000, 50000, length.out = 3))
  ),
  type = c(
    rep("host", 10),
    c(rep("host", 4), rep("prophage", 3), rep("host", 3))
  )
)
prophage_data$end <- prophage_data$start + 
  ifelse(prophage_data$type == "prophage", 4000, 3000)
prophage_data$orientation <- 1

cross_colors <- met.brewer("Cross", 5)
colors <- c("host" = cross_colors[3], "prophage" = cross_colors[1])

ggplot(prophage_data, aes(xmin = start, xmax = end, y = strain,
                          fill = type, forward = orientation, label = gene)) +
  geom_gene_arrow(arrowhead_height = unit(4, "mm"),
                  arrowhead_width = unit(2, "mm")) +
  geom_gene_label(align = "center", size = 2.5) +
  facet_wrap(~ strain, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = colors, 
                    labels = c("Host gene", "Prophage gene")) +
  labs(title = "Prophage Integration Event",
       subtitle = "Prophage inserted between Gene_4 and Gene_5 in lysogen",
       x = "Position (bp)", y = "", fill = "Gene Type") +
  theme_genes() +
  theme(legend.position = "bottom")
```

**Observation:** Prophage genes (red) inserted, disrupting original gene order
```{=html}
<!-- 
slide synteny7
-->
```

## Adding Gene Labels

**Use `geom_gene_label()` to add text:**
```{r}
#| echo: true
#| eval: false

ggplot(genes, aes(xmin = start, xmax = end, y = molecule,
                  fill = gene, forward = orientation,
                  label = gene)) +  # Add label aesthetic
  geom_gene_arrow() +
  geom_gene_label(align = "center") +  # Add labels!
  theme_genes() +
  theme(legend.position = "none")  # Remove legend
```

**Options for `geom_gene_label()`:**

- `align = "left"` / `"center"` / `"right"`: Label position
- `size = 3`: Text size
- `fontface = "bold"`: Make text bold
```{=html}
<!-- 
slide synteny8
-->
```

## Exercise 1: Gene Cluster Rearrangement

::: {.question}
**Create a biosynthetic gene cluster comparison:**
```{r}
#| echo: true
#| eval: false

# Two bacterial strains with different gene arrangements
cluster_data <- data.frame(
  strain = rep(c("Strain_1", "Strain_2"), each = 7),
  gene = rep(c("regulatorA", "enzymeB", "enzymeC", 
               "transporterD", "enzymeE", "enzymeF", "regulatorG"), 2),
  start = c(
    # Strain 1 (reference)
    c(1000, 3500, 6000, 8500, 11000, 13500, 16000),
    # Strain 2 (enzymeC deleted, transporterD moved to end)
    c(1000, 3500, 6000, 8500, 11000, 13500)
  ),
  type = c(
    rep(c("regulator", "enzyme", "enzyme", "transporter", 
          "enzyme", "enzyme", "regulator"), 2)
  )
)
# Add end positions
cluster_data$end <- cluster_data$start + 2000
cluster_data$orientation <- 1

# Your tasks:
# 1. Create gene arrow plot with facet_wrap by strain
# 2. Color by gene type (regulator, enzyme, transporter)
# 3. Add gene labels
# 4. Use Cross palette: regulators = color 1, enzymes = color 3, transporters = color 5
# 5. Add title describing the difference

# Hint: Create color vector like this:
# colors <- c("regulator" = cross_colors[1], 
#             "enzyme" = cross_colors[3], 
#             "transporter" = cross_colors[5])
```
:::
```{=html}
<!-- 
slide synteny9
-->
```

## Exercise 2: Prophage Insertion Analysis

::: {.question}
**Analyze prophage integration in E. coli:**
```{r}
#| echo: true
#| eval: false

# Wildtype vs prophage-containing strain
set.seed(123)

ecoli_prophage <- data.frame(
  strain = rep(c("E.coli_K12", "E.coli_Lysogen"), each = 12),
  gene = c(
    # Wildtype genes
    paste0("Gene_", sprintf("%02d", 1:12)),
    # Lysogen: prophage inserted after Gene_06
    c(paste0("Gene_", sprintf("%02d", 1:6)),
      paste0("λ_", c("CI", "Cro", "N", "Q", "Tail")),  # 5 phage genes
      paste0("Gene_", sprintf("%02d", 7:7)))
  ),
  start = c(
    # Wildtype: evenly spaced
    seq(10000, 110000, length.out = 12),
    # Lysogen: disrupted spacing
    c(seq(10000, 60000, length.out = 6),   # Genes 1-6
      seq(62000, 90000, length.out = 5),    # Prophage
      seq(92000, 92000, length.out = 1))    # Gene 7 (shifted)
  ),
  type = c(
    rep("host", 12),
    c(rep("host", 6), rep("prophage", 5), rep("host", 1))
  )
)
ecoli_prophage$end <- ecoli_prophage$start + 
  ifelse(ecoli_prophage$type == "prophage", 5000, 7000)
ecoli_prophage$orientation <- sample(c(1, -1), nrow(ecoli_prophage), 
                                     replace = TRUE)

# Your tasks:
# 1. Plot both strains with facet_wrap
# 2. Color host genes vs prophage genes differently
# 3. Add gene labels (use size = 2 for readability)
# 4. Add arrows showing gene orientation
# 5. Highlight the insertion site with annotation

# Bonus: Calculate total genome size increase due to prophage
```
:::
```{=html}
<!-- 
slide synteny10
-->
```

## Key Points: gggenes

**Remember:**

✅ **Data structure:** molecule, gene, start, end, orientation  
✅ **Aesthetics:** `xmin`, `xmax`, `y`, `forward`  
✅ **Arrow direction:** `orientation = 1` (→) or `-1` (←)  
✅ **Faceting:** Use `facet_wrap(~ molecule)` for comparisons  
✅ **Colors:** Map to gene type or identity  
✅ **Labels:** Add with `geom_gene_label()`

**Next:** Large-scale synteny with syntenyPlotteR
